---
title: "Projected Temperature-Attributable Mortality for Madrid"
subtitle: "Part 1: Attributable Numbers by Age Group and Temperature Range"
author: "Daniel Precioso"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    fig_width: 10
    fig_height: 6
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    keep_tex: false
---

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = FALSE,
  comment = "##",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook demonstrates the first methodological block of the pipeline for a **single city** and a **single climate model**. It follows the standard attribution framework used in the temperature-mortality literature and stops at the calculation of **annual attributable numbers (ANs)** and **attributable fractions (AFs)**.

The notebook is deliberately restricted to:

- **one city**: Madrid
- **one SSP**: SSP3-7.0
- **one climate model**: selected from the available projected temperature dataset
- **no adaptation**
- **wide age groups** as in the Masselot framework:
  - 20–44
  - 45–64
  - 65–74
  - 75–84
  - 85+

This notebook does **not** construct life tables or compute life expectancy or lifespan inequality. Those steps belong to later notebooks in the series.

## Methodological aim

For each year and age group, we compute annual temperature-attributable mortality separately for four standard temperature ranges:

- **extreme cold**
- **moderate cold**
- **moderate heat**
- **extreme heat**

The calculation follows the standard attributable-risk logic described by Gasparrini and Leone (2014), using age-group-specific temperature-mortality exposure-response functions (ERFs) and projected daily temperatures.

## Main output

The main output is an annual tidy table with one row per:

- city
- climate model
- SSP
- year
- age group
- temperature range

containing:

- annual attributable number (**AN**)
- annual attributable fraction (**AF**)

---

# 1. Setup and Configuration

## 1.1 Load Required Libraries

```{r setup-libraries}
library(data.table)
library(arrow)
library(dlnm)
library(splines)
library(ggplot2)
library(scales)

cat("Libraries loaded successfully.\n")
```

---

## 1.2 Define Demo Configuration

```{r set-config}
# Demo city
city <- "Madrid"
city_code <- "ES001C"

# Scenario selection
scenario_label <- "SSP3-7.0"
ssp_value <- 3

# Climate model selection
# If demo_gcm is NA, the notebook will automatically use the first available GCM.
demo_gcm <- NA_character_

# No adaptation in this notebook
adaptation_level <- 0

# Time windows
historical_year_min <- 1990
historical_year_max <- 2019
projection_year_min <- 2020
projection_year_max <- 2100

# Standard temperature-range thresholds
cold_extreme_percentile <- 0.025
heat_extreme_percentile <- 0.975

# Wide age groups as used in Masselot-style prediction
age_groups <- c("20-44", "45-64", "65-74", "75-84", "85+")

# Input data paths
coefs_path <- "../data/coefs.csv"
temp_path <- "../data/tmeanproj.gz.parquet"
demography_path <- file.path(tempdir(), "city_demography.csv")

# Output paths — single unified directory
output_dir <- "../results/demo"
city_prefix <- tolower(gsub(" ", "_", city))
output_file_annual <- file.path(output_dir, paste0(city_prefix, "_ans_annual.csv"))
output_file_daily <- file.path(output_dir, paste0(city_prefix, "_ans_daily.csv"))

cat("=== Demo Configuration ===\n")
cat(sprintf("  City: %s (%s)\n", city, city_code))
cat(sprintf("  Scenario: %s\n", scenario_label))
cat(sprintf("  Adaptation: %.0f%%\n", adaptation_level * 100))
cat(sprintf("  Historical reference years: %d–%d\n", historical_year_min, historical_year_max))
cat(sprintf("  Projection years: %d–%d\n", projection_year_min, projection_year_max))
cat(sprintf("  Age groups: %s\n", paste(age_groups, collapse = ", ")))
```

---

# 2. Data Requirements and Assumptions

This notebook assumes the following data are already available in the project folder:

1. **City-specific ERF coefficients** for Madrid in the wide age groups.
2. **Historical and projected daily temperatures** for Madrid, including multiple GCMs and SSPs.
3. **Annual demographic baseline inputs** for Madrid by age group and year, including either:
   - annual deaths directly, or
   - population and crude mortality rate from which annual deaths can be derived.

## 2.1 Expected Demography Table Structure

The notebook assumes `demography_path` contains at least these columns:

- `city_code`
- `year`
- `age_group`
- `population`
- and either:
  - `annual_deaths`, or
  - `mx`

If `annual_deaths` is missing but `mx` is available, the notebook derives:

$$
\text{annual deaths} = m_x \times \text{population}
$$

## 2.2 Important methodological notes

- This notebook uses the **wide age groups** directly and does **not** yet disaggregate to single years of age.
- It uses **no adaptation**.
- It computes **annual ANs and AFs** only.
- It assumes that annual expected deaths can be distributed evenly across days within each year as an approximation for impact assessment when city-wide daily deaths are not available for the future projection workflow.
- This is consistent with the style of impact assessment used in the projection literature when projected city-wide daily mortality series are not directly available.

---

# 3. Load Input Data

## 3.1 Helper Functions

```{r helper-functions}
check_required_cols <- function(dt, cols, table_name) {
  missing_cols <- setdiff(cols, names(dt))
  if (length(missing_cols) > 0L) {
    stop(table_name, " is missing required columns: ", paste(missing_cols, collapse = ", "))
  }
}

safe_id <- function(x) {
  gsub("[^A-Za-z0-9_]+", "_", x)
}

assign_temp_range <- function(temp, p025, p975, mmt) {
  fifelse(
    temp < p025, "extreme_cold",
    fifelse(
      temp < mmt, "moderate_cold",
      fifelse(
        temp <= p975, "moderate_heat",
        "extreme_heat"
      )
    )
  )
}

compute_centered_rr <- function(temp, coef_vec, argvar, center_value) {
  basis_temp <- do.call(onebasis, c(list(x = temp), argvar))
  basis_center <- do.call(onebasis, c(list(x = rep(center_value, length(temp))), argvar))
  coef_mat <- matrix(coef_vec, nrow = length(temp), ncol = length(coef_vec), byrow = TRUE)
  log_rr <- rowSums((basis_temp - basis_center) * coef_mat)
  pmax(exp(log_rr), 1)
}
```

---

## 3.2 Load ERF Coefficients

```{r load-coefficients}
coefs <- fread(coefs_path)

required_coef_cols <- c("URAU_CODE", "agegroup", "b1", "b2", "b3", "b4", "b5")
check_required_cols(coefs, required_coef_cols, "Coefficient table")

coefs_city <- coefs[URAU_CODE == city_code]

if (nrow(coefs_city) == 0L) {
  stop("No ERF coefficients found for city_code=", city_code, " or city=", city)
}

coefs_city <- coefs_city[agegroup %in% age_groups]
# setorder(coefs_city, match(agegroup, age_groups))

if (nrow(coefs_city) != length(age_groups)) {
  stop("ERF coefficient table does not contain all expected age groups for Madrid.")
}

cat("=== ERF Coefficients Loaded ===\n")
print(coefs_city[, .(city, city_code, agegroup, b1, b2, b3, b4, b5)])
```

---

## 3.3 Load Daily Temperature Data

```{r load-temperature}
temp_all <- as.data.table(read_parquet(temp_path))

# Minimal required columns straight from parquet
check_required_cols(temp_all, c("URAU_CODE", "date", "ssp"), "Temperature table")

# Ensure date is a proper Date and derive year/doy if missing
if (!inherits(temp_all$date, "Date")) {
  temp_all[, date := as.Date(date)]
}
if (anyNA(temp_all$date)) {
  stop("Temperature table contains invalid or missing date values.")
}
if (!"year" %in% names(temp_all)) {
  temp_all[, year := as.integer(format(date, "%Y"))]
}
if (!"doy" %in% names(temp_all)) {
  temp_all[, doy := as.integer(format(date, "%j"))]
}

required_temp_cols <- c("URAU_CODE", "date", "year", "doy", "ssp")
check_required_cols(temp_all, required_temp_cols, "Temperature table after deriving calendar fields")

temp_city <- temp_all[URAU_CODE == city_code]

if (nrow(temp_city) == 0L) {
  stop("No temperature rows found for city_code=", city_code)
}

# Identify GCM columns as all non-id columns except the expected index columns
id_cols <- c("URAU_CODE", "date", "year", "doy", "ssp")
gcm_cols <- setdiff(names(temp_city), id_cols)

if (length(gcm_cols) == 0L) {
  stop("No GCM columns detected in the temperature table.")
}

if (is.na(demo_gcm)) {
  demo_gcm <- gcm_cols[[1]]
}

if (!demo_gcm %in% gcm_cols) {
  stop("Selected demo_gcm not found in projected temperature table: ", demo_gcm)
}

hist_temp <- temp_city[
  ssp == "hist" & year >= historical_year_min & year <= historical_year_max,
  .(date, year, doy, temp = get(demo_gcm))
]
proj_temp <- temp_city[
  ssp == ssp_value & year >= projection_year_min & year <= projection_year_max,
  .(date, year, doy, temp = get(demo_gcm))
]

if (nrow(hist_temp) == 0L) {
  stop("No historical daily temperature data found for the selected city/model.")
}
if (nrow(proj_temp) == 0L) {
  stop("No projected daily temperature data found for the selected city/model and SSP.")
}
if (anyNA(hist_temp$temp)) {
  stop("Historical temperature series contains missing values for the selected model.")
}
if (anyNA(proj_temp$temp)) {
  stop("Projected temperature series contains missing values for the selected model.")
}

cat("=== Temperature Data Loaded ===\n")
cat(sprintf("  City: %s\n", city_code))
cat(sprintf("  Selected GCM: %s\n", demo_gcm))
cat(sprintf("  Historical rows: %s\n", format(nrow(hist_temp), big.mark = ",")))
cat(sprintf("  Projected rows: %s\n", format(nrow(proj_temp), big.mark = ",")))
cat(sprintf("  Historical years: %d–%d\n", min(hist_temp$year), max(hist_temp$year)))
cat(sprintf("  Projected years: %d–%d\n", min(proj_temp$year), max(proj_temp$year)))
```

---

## 3.3.5 Build Demographic Baseline Table

The demographic table (`city_demography.csv`) is not stored as source data — it is derived from the `city_results.csv` baseline (population and deaths by age group for Madrid), with death rates held constant across projection years. This is appropriate for a methodological demo focused on temperature attribution.

```{r build-demography}
city_res <- fread("../data/city_results.csv")
madrid_base <- city_res[URAU_CODE == city_code]
madrid_base <- madrid_base[agegroup %in% age_groups]

base_pop  <- setNames(madrid_base$agepop, madrid_base$agegroup)
base_death <- setNames(madrid_base$death, madrid_base$agegroup)
base_mx   <- base_death / base_pop

all_years <- (projection_year_min - 1L):projection_year_max
demo_city <- rbindlist(lapply(age_groups, function(ag) {
  data.table(
    city_code = city_code,
    year = all_years,
    age_group = ag,
    population = round(base_pop[ag]),
    mx = base_mx[ag],
    annual_deaths = round(base_mx[ag] * base_pop[ag])
  )
}))
setcolorder(demo_city, c("city_code", "year", "age_group", "population", "annual_deaths", "mx"))

fwrite(demo_city, demography_path)

cat("=== Demographic Baseline Built ===\n")
cat(sprintf("  Rows: %s\n", format(nrow(demo_city), big.mark = ",")))
cat(sprintf("  Years: %d–%d\n", min(demo_city$year), max(demo_city$year)))
```

---

## 3.4 Load Demographic Baseline Inputs

```{r load-demography}
demo <- fread(demography_path)
check_required_cols(demo, c("city_code", "year", "age_group", "population"), "Demography table")

cc <- city_code
demo_city <- demo[city_code == cc & year >= projection_year_min & year <= projection_year_max]

if (nrow(demo_city) == 0L) {
  stop("No demographic rows found for Madrid in the projection years.")
}

if (!"annual_deaths" %in% names(demo_city)) {
  if (!"mx" %in% names(demo_city)) {
    stop("Demography table must contain either annual_deaths or mx.")
  }
  demo_city[, annual_deaths := mx * population]
}

check_required_cols(demo_city, c("year", "age_group", "population", "annual_deaths"), "Demography table after derivation")

demo_city <- demo_city[age_group %in% age_groups]
demo_city <- demo_city[order(year, match(age_group, age_groups))]

if (anyNA(demo_city$annual_deaths) || any(demo_city$annual_deaths < 0)) {
  stop("Annual deaths contain missing or negative values.")
}

cat("=== Demographic Baseline Loaded ===\n")
cat(sprintf("  Rows: %s\n", format(nrow(demo_city), big.mark = ",")))
cat(sprintf("  Years: %d–%d\n", min(demo_city$year), max(demo_city$year)))
print(head(demo_city, 10))
```

---

# 4. Reconstruct Madrid ERFs for the Wide Age Groups

The Masselot framework estimates age-group-specific cumulative temperature-mortality relationships using spline-based exposure-response functions. In this notebook, we reconstruct those wide-age-group ERFs for Madrid using the coefficient vector for each age group and a city-specific temperature basis built from the historical temperature distribution.

## 4.1 Define the Temperature Basis

```{r define-basis}
varfun <- "bs"
vardegree <- 2
varper <- c(10, 75, 90)

varknots <- quantile(hist_temp$temp, probs = varper / 100, na.rm = TRUE)
varbound <- range(hist_temp$temp, na.rm = TRUE)

argvar <- list(
  fun = varfun,
  degree = vardegree,
  knots = varknots,
  Bound = varbound
)

cat("=== Temperature Basis Parameters ===\n")
cat(sprintf("  Bounds: [%.2f, %.2f]\n", varbound[1], varbound[2]))
cat(sprintf("  Knots (10th, 75th, 90th pct): %s\n", paste(round(varknots, 2), collapse = ", ")))
```

---

## 4.2 Derive Age-Group-Specific MMTs

For each age group, we evaluate the cumulative ERF over a fine temperature grid and identify the minimum mortality temperature (MMT). Following common practice, the search is restricted to the central-to-upper part of the local temperature distribution to avoid unstable minima at the far extremes.

```{r derive-mmt}
temp_seq <- seq(varbound[1], varbound[2], by = 0.1)

rr_curves <- rbindlist(lapply(seq_len(nrow(coefs_city)), function(i) {
  ag <- coefs_city$agegroup[[i]]
  coef_vec <- as.numeric(coefs_city[i, .(b1, b2, b3, b4, b5)])

  basis_seq <- do.call(onebasis, c(list(x = temp_seq), argvar))
  log_rr_seq <- as.numeric(basis_seq %*% coef_vec)

  search_window <- temp_seq >= quantile(hist_temp$temp, 0.25, na.rm = TRUE) &
                   temp_seq <= quantile(hist_temp$temp, 0.99, na.rm = TRUE)

  mmt_value <- temp_seq[search_window][which.min(log_rr_seq[search_window])]

  rr_centered <- compute_centered_rr(temp_seq, coef_vec, argvar, mmt_value)

  data.table(
    age_group = ag,
    temp = temp_seq,
    rr = rr_centered,
    mmt = mmt_value
  )
}))

mmt_by_age <- unique(rr_curves[, .(age_group, mmt)])

cat("=== Derived MMT by Age Group ===\n")
print(mmt_by_age)
```

---

## 4.3 Plot the Reconstructed ERFs

```{r plot-erfs}
ggplot(rr_curves, aes(x = temp, y = rr, color = age_group)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, color = "grey40", linewidth = 0.5) +
  geom_vline(data = mmt_by_age, aes(xintercept = mmt, color = age_group),
             linetype = "dashed", alpha = 0.8, show.legend = FALSE) +
  labs(
    title = sprintf("Reconstructed Temperature-Mortality ERFs — %s", city),
    subtitle = sprintf("Wide age groups | %s | GCM: %s", scenario_label, demo_gcm),
    x = "Daily mean temperature (°C)",
    y = "Relative risk (RR)",
    color = "Age group"
  ) +
  theme_minimal(base_size = 12)
```

---

# 5. Define Standard Temperature Ranges

We now classify each projected daily temperature into one of four standard ranges:

- **extreme cold**: below the 2.5th percentile of the historical city distribution
- **moderate cold**: between the 2.5th percentile and the age-group-specific MMT
- **moderate heat**: between the age-group-specific MMT and the 97.5th percentile of the historical city distribution
- **extreme heat**: above the 97.5th percentile of the historical city distribution

## 5.1 Compute Historical Reference Thresholds

```{r define-thresholds}
p025_hist <- as.numeric(quantile(hist_temp$temp, cold_extreme_percentile, na.rm = TRUE))
p975_hist <- as.numeric(quantile(hist_temp$temp, heat_extreme_percentile, na.rm = TRUE))

cat("=== Historical Thresholds for Range Classification ===\n")
cat(sprintf("  2.5th percentile: %.2f °C\n", p025_hist))
cat(sprintf("  97.5th percentile: %.2f °C\n", p975_hist))
```

---

# 6. Compute Daily RR, AF, and AN

The attributable-risk calculation is based on the standard epidemiological identity:

$$
AF = \frac{RR - 1}{RR}
$$

For each projected day and age group, we compute:

1. temperature-specific RR relative to the age-group-specific MMT,
2. the corresponding AF,
3. the attributable number of deaths:

$$
AN_{day} = AF_{day} \times \frac{D_{year}}{N_{days,year}}
$$

where $$D_{year}$$ is the annual expected number of deaths in that age group and year.

## 6.1 Expand the Daily Projection Data Across Age Groups

```{r expand-projection}
proj_temp[, date := as.Date(date)]
proj_temp[, n_days_year := .N, by = year]

proj_long <- CJ(
  date = proj_temp$date,
  age_group = age_groups,
  unique = TRUE
)
proj_long <- merge(proj_long, proj_temp, by = "date", all.x = TRUE)
proj_long <- merge(proj_long, mmt_by_age, by = "age_group", all.x = TRUE)
proj_long <- merge(proj_long, demo_city[, .(year, age_group, annual_deaths)],
                   by = c("year", "age_group"), all.x = TRUE)

if (anyNA(proj_long$temp) || anyNA(proj_long$mmt) || anyNA(proj_long$annual_deaths)) {
  stop("Projected long table contains missing temperature, MMT, or annual death values.")
}

cat("=== Expanded Projected Table ===\n")
cat(sprintf("  Rows: %s\n", format(nrow(proj_long), big.mark = ",")))
print(head(proj_long, 10))
```

---

## 6.2 Compute Daily RR and AF

```{r compute-daily-rr-af}
coef_lookup <- coefs_city[, .(agegroup, b1, b2, b3, b4, b5)]
setnames(coef_lookup, "agegroup", "age_group")
proj_long <- merge(proj_long, coef_lookup, by = "age_group", all.x = TRUE)

proj_long[, rr := compute_centered_rr(
  temp = temp,
  coef_vec = c(b1[1], b2[1], b3[1], b4[1], b5[1]),
  argvar = argvar,
  center_value = mmt[1]
), by = .(age_group)]

proj_long[, af_total := (rr - 1) / rr]
proj_long[af_total < 0, af_total := 0]

proj_long[, temp_range := assign_temp_range(temp, p025_hist, p975_hist, mmt)]

# Attribute AF only to the daily range in which the temperature falls
proj_long[, af_range := af_total]
proj_long[, daily_expected_deaths := annual_deaths / n_days_year]
proj_long[, an_day := af_range * daily_expected_deaths]

cat("=== Daily RR/AF Computed ===\n")
print(head(proj_long[, .(date, year, age_group, temp, mmt, rr, af_total, temp_range, an_day)], 12))
```

---

# 7. Aggregate to Annual ANs and AFs

## 7.1 Annual ANs by Age Group and Temperature Range

```{r annual-ans}
annual_ans <- proj_long[, .(
  AN = sum(an_day, na.rm = TRUE),
  annual_deaths = unique(annual_deaths),
  mean_rr = mean(rr, na.rm = TRUE),
  mean_af = mean(af_total, na.rm = TRUE)
), by = .(year, age_group, temp_range)]

annual_ans[, AF := AN / annual_deaths]
annual_ans[, `:=`(
  city = city,
  city_code = city_code,
  gcm = demo_gcm,
  ssp = scenario_label,
  adaptation = adaptation_level
)]

setcolorder(annual_ans, c(
  "city", "city_code", "gcm", "ssp", "adaptation",
  "year", "age_group", "temp_range", "annual_deaths", "AN", "AF", "mean_rr", "mean_af"
))

cat("=== Annual AN Table ===\n")
print(head(annual_ans, 20))
```

---

## 7.2 Annual Total Across Temperature Ranges

```{r annual-totals}
annual_totals <- annual_ans[, .(
  annual_deaths = unique(annual_deaths),
  AN_total = sum(AN),
  AF_total = sum(AN) / unique(annual_deaths)
), by = .(year, age_group)]

cat("=== Annual Totals Across Temperature Ranges ===\n")
print(head(annual_totals, 20))
```

---

# 8. Diagnostics and Interpretation

## 8.1 Annual Attributable Numbers by Temperature Range

```{r plot-annual-ans-range}
ggplot(annual_ans, aes(x = year, y = AN, color = temp_range)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ age_group, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(
    title = sprintf("Projected Annual Attributable Numbers — %s", city),
    subtitle = sprintf("%s | %s | No adaptation | Wide age groups", scenario_label, demo_gcm),
    x = "Year",
    y = "Attributable deaths (AN)",
    color = "Temperature range"
  ) +
  theme_minimal(base_size = 12)
```

---

## 8.2 Annual Attributable Fraction by Temperature Range

```{r plot-annual-af-range}
ggplot(annual_ans, aes(x = year, y = AF, color = temp_range)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ age_group, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = sprintf("Projected Annual Attributable Fractions — %s", city),
    subtitle = sprintf("%s | %s | No adaptation | Wide age groups", scenario_label, demo_gcm),
    x = "Year",
    y = "Attributable fraction",
    color = "Temperature range"
  ) +
  theme_minimal(base_size = 12)
```

---

## 8.3 Age Pattern in a Selected Year

```{r age-pattern-single-year}
selected_year <- min(annual_ans$year)

age_plot_dt <- annual_ans[year == selected_year]

ggplot(age_plot_dt, aes(x = age_group, y = AN, fill = temp_range)) +
  geom_col(position = "stack") +
  scale_y_continuous(labels = comma) +
  labs(
    title = sprintf("Projected Attributable Numbers by Age Group — %s", selected_year),
    subtitle = sprintf("%s | %s | %s", city, scenario_label, demo_gcm),
    x = "Age group",
    y = "Attributable deaths (AN)",
    fill = "Temperature range"
  ) +
  theme_minimal(base_size = 12)
```

---

# 9. Export Results

```{r export-results}
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

fwrite(
  annual_ans,
  output_file_annual
)

fwrite(
  proj_long[, .(
    city = city,
    city_code = city_code,
    gcm = demo_gcm,
    ssp = scenario_label,
    adaptation = adaptation_level,
    date,
    year,
    doy,
    age_group,
    temp,
    mmt,
    rr,
    af_total,
    temp_range,
    daily_expected_deaths,
    an_day
  )],
  output_file_daily
)

cat("=== Files Written ===\n")
cat(sprintf("  Annual AN table: %s\n", output_file_annual))
cat(sprintf("  Daily attribution table: %s\n", output_file_daily))
```

---

# 10. Notebook Summary

This notebook has:

1. loaded city-specific wide-age-group ERFs for Madrid,
2. reconstructed temperature-mortality RR curves using the standard spline basis,
3. derived age-group-specific MMTs,
4. classified projected daily temperatures into four standard ranges,
5. computed daily RR, AF, and attributable numbers,
6. aggregated those results into annual ANs and AFs by age group and temperature range.

## Main deliverable for the next notebook

The main output is the annual AN table by:

- year
- age group
- temperature range

The next notebook in the series will take this output and **disaggregate the age-group ANs into single years of age using `pclm`**.

---

# Session Information

```{r session-info}
sessionInfo()
```