---
title: "Projected Temperature-Attributable Mortality for Madrid"
subtitle: "Part 1: Attributable Numbers by Age Group and Temperature Range"
author: "Daniel Precioso"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = FALSE,
  comment = "##",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook demonstrates the first methodological block of the pipeline for a **single city** and a **single climate model**. It follows the standard attribution framework used in the temperature-mortality literature (Gasparrini & Leone 2014, Masselot et al. 2023) and stops at the calculation of **annual attributable numbers (ANs)** and **attributable fractions (AFs)**.

The notebook is deliberately restricted to:

- **one city**: Madrid
- **one SSP**: SSP3-7.0
- **one climate model**: selected from the available projected temperature dataset
- **no adaptation**
- **wide age groups** as in the Masselot framework:
  - 20–44
  - 45–64
  - 65–74
  - 75–84
  - 85+

This notebook does **not** construct life tables or compute life expectancy or lifespan inequality. Those steps belong to later notebooks in the series.

## Methodological aim

For each year and age group, we compute annual temperature-attributable mortality separately for four standard temperature ranges:

- **extreme cold**
- **moderate cold**
- **moderate heat**
- **extreme heat**

The calculation follows the standard attributable-risk logic described by Gasparrini and Leone (2014), using age-group-specific temperature-mortality exposure-response functions (ERFs) and projected daily temperatures.

## Main output

The main output is an annual tidy table with one row per:

- city
- climate model
- SSP
- year
- age group
- temperature range

containing:

- annual attributable number (**AN**) with 95% CI
- annual attributable fraction (**AF**)

---

# Setup and Configuration

## Load Required Libraries

```{r setup-libraries}
library(data.table)
library(arrow)
library(dlnm)
library(splines)
library(ggplot2)
library(scales)
library(MASS)

cat("Libraries loaded successfully.\n")
```

## Source Project R Functions

```{r source-helpers}
source("../R/rr_basis.R")
source("../R/simulation.R")
source("../R/utils.R")
```

---

## Define Demo Configuration

```{r set-config}
# Demo city
city <- "Madrid"
city_code <- "ES001C"

# Scenario selection
scenario_label <- "SSP3-7.0"
ssp_value <- 3

# Climate model selection
# If demo_gcm is NA, the notebook will automatically use the first available GCM.
demo_gcm <- NA_character_

# No adaptation in this notebook
adaptation_level <- 0

# Time windows
historical_year_min <- 1990
historical_year_max <- 2019
projection_year_min <- 2020
projection_year_max <- 2100

# Standard temperature-range thresholds
cold_extreme_percentile <- 0.025
heat_extreme_percentile <- 0.975

# Wide age groups as used in Masselot-style prediction
age_groups <- c("20-44", "45-64", "65-74", "75-84", "85+")

# Basis parameters (matching Masselot 2023)
varfun <- "bs"
vardegree <- 2
varper <- c(10, 75, 90)

# Input data paths
coefs_path <- "../data/coefs.csv"
simu_path <- "../data/coef_simu.csv"
temp_path <- "../data/tmeanproj.gz.parquet"
demography_path <- file.path(tempdir(), "city_demography.csv")

# Output paths — single unified directory
output_dir <- "../results/demo"
city_prefix <- tolower(gsub(" ", "_", city))
output_file_annual <- file.path(output_dir, paste0(city_prefix, "_ans_annual.csv"))
output_file_daily <- file.path(output_dir, paste0(city_prefix, "_ans_daily.csv"))

cat("=== Demo Configuration ===\n")
cat(sprintf("  City: %s (%s)\n", city, city_code))
cat(sprintf("  Scenario: %s\n", scenario_label))
cat(sprintf("  Adaptation: %.0f%%\n", adaptation_level * 100))
cat(sprintf("  Historical reference years: %d–%d\n", historical_year_min, historical_year_max))
cat(sprintf("  Projection years: %d–%d\n", projection_year_min, projection_year_max))
cat(sprintf("  Age groups: %s\n", paste(age_groups, collapse = ", ")))
```

---

# Data Requirements and Assumptions

This notebook assumes the following data are already available in the project folder:

1. **City-specific ERF coefficients** for Madrid in the wide age groups (`data/coefs.csv`).
2. **Pre-simulated coefficients** for CI computation (`data/coef_simu.csv`).
3. **Historical and projected daily temperatures** for Madrid, including multiple GCMs and SSPs (`data/tmeanproj.gz.parquet`).
4. **Annual demographic baseline inputs** for Madrid by age group and year (`data/city_results.csv`).

## Important methodological notes

- This notebook uses the **wide age groups** directly and does **not** yet disaggregate to single years of age.
- It uses **no adaptation**.
- It computes **annual ANs and AFs** with 95% confidence intervals via simulation (nsim=1000), following the Masselot approach.
- Confidence intervals use pre-simulated coefficients from the original Masselot second-stage meta-regression.
- Annual expected deaths are distributed evenly across days within each year.

---

# Load Input Data

## Helper Functions

```{r helper-functions}
check_required_cols <- function(dt, cols, table_name) {
  missing_cols <- setdiff(cols, names(dt))
  if (length(missing_cols) > 0L)
    stop(table_name, " is missing required columns: ", paste(missing_cols, collapse = ", "))
}
```

---

## Load ERF Coefficients

```{r load-coefficients}
coefs <- fread(coefs_path)

required_coef_cols <- c("URAU_CODE", "agegroup", "b1", "b2", "b3", "b4", "b5")
check_required_cols(coefs, required_coef_cols, "Coefficient table")

coefs_city <- coefs[URAU_CODE == city_code]

if (nrow(coefs_city) == 0L)
  stop("No ERF coefficients found for city_code=", city_code)

coefs_city <- coefs_city[agegroup %in% age_groups]

if (nrow(coefs_city) != length(age_groups))
  stop("ERF coefficient table does not contain all expected age groups.")

cat("=== ERF Coefficients Loaded ===\n")
print(coefs_city[, .(URAU_CODE, agegroup, b1, b2, b3, b4, b5)])
```

## Load Pre-Simulated Coefficients (for CIs)

```{r load-simulated-coefs}
# Load pre-simulated coefficients from Masselot second-stage (nsim=1000)
coef_simu <- fread(simu_path)
coef_simu_city <- coef_simu[URAU_CODE == city_code & agegroup %in% age_groups]

cat("=== Pre-Simulated Coefficients ===\n")
cat(sprintf("  Rows (city x agegroup x sim): %s\n", format(nrow(coef_simu_city), big.mark = ",")))
cat(sprintf("  Simulations per age group: %d\n", uniqueN(coef_simu_city$sim)))
```

---

## Load Daily Temperature Data

```{r load-temperature}
temp_all <- as.data.table(read_parquet(temp_path))

check_required_cols(temp_all, c("URAU_CODE", "date", "ssp"), "Temperature table")

if (!inherits(temp_all$date, "Date"))
  temp_all[, date := as.Date(date)]
if (!"year" %in% names(temp_all))
  temp_all[, year := as.integer(format(date, "%Y"))]
if (!"doy" %in% names(temp_all))
  temp_all[, doy := as.integer(format(date, "%j"))]

temp_city <- temp_all[URAU_CODE == city_code]

if (nrow(temp_city) == 0L)
  stop("No temperature rows found for city_code=", city_code)

id_cols <- c("URAU_CODE", "date", "year", "doy", "ssp")
gcm_cols <- setdiff(names(temp_city), id_cols)

if (length(gcm_cols) == 0L)
  stop("No GCM columns detected in the temperature table.")

if (is.na(demo_gcm))
  demo_gcm <- gcm_cols[[1]]

if (!demo_gcm %in% gcm_cols)
  stop("Selected demo_gcm not found: ", demo_gcm)

hist_temp <- temp_city[
  ssp == "hist" & year >= historical_year_min & year <= historical_year_max,
  .(date, year, doy, temp = get(demo_gcm))]
proj_temp <- temp_city[
  ssp == ssp_value & year >= projection_year_min & year <= projection_year_max,
  .(date, year, doy, temp = get(demo_gcm))]

if (nrow(hist_temp) == 0L)
  stop("No historical temperature data found.")
if (nrow(proj_temp) == 0L)
  stop("No projected temperature data found.")

cat("=== Temperature Data Loaded ===\n")
cat(sprintf("  City: %s | GCM: %s\n", city_code, demo_gcm))
cat(sprintf("  Historical: %s rows (%d–%d)\n",
  format(nrow(hist_temp), big.mark = ","),
  min(hist_temp$year), max(hist_temp$year)))
cat(sprintf("  Projected: %s rows (%d–%d)\n",
  format(nrow(proj_temp), big.mark = ","),
  min(proj_temp$year), max(proj_temp$year)))
```

---

## Build Demographic Baseline

```{r build-demography}
city_res <- fread("../data/city_results.csv")
madrid_base <- city_res[URAU_CODE == city_code]
madrid_base <- madrid_base[agegroup %in% age_groups]

base_pop <- setNames(madrid_base$agepop, madrid_base$agegroup)
base_death <- setNames(madrid_base$death, madrid_base$agegroup)
base_mx <- base_death / base_pop

all_years <- (projection_year_min - 1L):projection_year_max
demo_city <- rbindlist(lapply(age_groups, function(ag) {
  data.table(
    city_code = city_code,
    year = all_years,
    age_group = ag,
    population = round(base_pop[ag]),
    mx = base_mx[ag],
    annual_deaths = round(base_mx[ag] * base_pop[ag]))
}))
setcolorder(demo_city, c("city_code", "year", "age_group", "population", "annual_deaths", "mx"))

fwrite(demo_city, demography_path)

demo <- fread(demography_path)
demo_city <- demo[city_code == city_code & year >= projection_year_min & year <= projection_year_max]
demo_city <- demo_city[age_group %in% age_groups]
demo_city <- demo_city[order(year, match(age_group, age_groups))]

cat("=== Demographic Baseline ===\n")
cat(sprintf("  Years: %d–%d | Rows: %s\n",
  min(demo_city$year), max(demo_city$year),
  format(nrow(demo_city), big.mark = ",")))
```

---

# Reconstruct Madrid ERFs (Masselot-style)

The Masselot framework estimates age-group-specific cumulative temperature-mortality relationships using B-spline exposure-response functions (knots at 10th, 75th, 90th percentiles, degree 2). We reconstruct these from the pre-estimated coefficients.

## Build Temperature Basis

```{r define-basis}
basis_params <- build_basis_params(hist_temp$temp, varfun, vardegree, varper)
argvar <- basis_params$argvar
varbound <- basis_params$varbound
varknots <- basis_params$varknots

cat("=== Temperature Basis Parameters ===\n")
cat(sprintf("  %s (degree=%d), knots at %s pctiles\n",
  varfun, vardegree, paste(varper, collapse = ", ")))
cat(sprintf("  Knot values: %s\n", paste(round(varknots, 2), collapse = ", ")))
cat(sprintf("  Bounds: [%.2f, %.2f]°C\n", varbound[1], varbound[2]))
```

## Compute RR Curves and MMTs

```{r derive-mmt}
rr_curves_result <- compute_rr_curves(coefs_city, age_groups, argvar, varbound)
rr_curves <- data.table(
  temp = rep(rr_curves_result$temp_seq, length(age_groups)),
  age_group = rep(age_groups, each = length(rr_curves_result$temp_seq)),
  rr = as.vector(rr_curves_result$rr_matrix))

mmt_by_age <- data.table(
  age_group = age_groups,
  mmt = rr_curves_result$mmt_vec)

cat("=== Derived MMT by Age Group ===\n")
print(mmt_by_age)
```

---

# Define Temperature Ranges

Each projected daily temperature is classified into one of four ranges based on the age-group-specific MMT and historical percentiles.

```{r define-thresholds}
p025_hist <- as.numeric(quantile(hist_temp$temp, cold_extreme_percentile, na.rm = TRUE))
p975_hist <- as.numeric(quantile(hist_temp$temp, heat_extreme_percentile, na.rm = TRUE))

cat("=== Historical Thresholds ===\n")
cat(sprintf("  2.5th pctile: %.2f°C | 97.5th pctile: %.2f°C\n", p025_hist, p975_hist))
```

---

# Compute Daily RR, AF, and AN (Masselot-style)

We use the standard attributable-risk identity:

$$AF = 1 - e^{-X\beta}, \quad AN = AF \times \text{daily deaths}$$

where $X\beta$ is the centred B-spline basis evaluated at daily temperature.

```{r expand-projection}
proj_temp[, date := as.Date(date)]
proj_temp[, n_days_year := .N, by = year]

proj_long <- CJ(date = proj_temp$date, age_group = age_groups, unique = TRUE)
proj_long <- merge(proj_long, proj_temp, by = "date", all.x = TRUE)
proj_long <- merge(proj_long, mmt_by_age, by = "age_group", all.x = TRUE)
proj_long <- merge(proj_long, demo_city[, .(year, age_group, annual_deaths)],
  by = c("year", "age_group"), all.x = TRUE)

cat("=== Projected Table Expanded ===\n")
cat(sprintf("  Rows: %s\n", format(nrow(proj_long), big.mark = ",")))
```

```{r compute-daily-rr-af-masselot}
# Masselot-style: onebasis -> center -> 1 - exp(-bvarcen %*% coef)
coef_lookup <- coefs_city[, .(agegroup, b1, b2, b3, b4, b5)]
setnames(coef_lookup, "agegroup", "age_group")
proj_long <- merge(proj_long, coef_lookup, by = "age_group", all.x = TRUE)

proj_long[, c("af_day", "an_day") := {
  bvar <- do.call(onebasis, c(list(x = temp), argvar))
  cenvec <- do.call(onebasis,
    c(list(x = mmt[1]), argvar))
  bvarcen <- scale(bvar, center = cenvec, scale = FALSE)
  coef_vec <- c(b1[1], b2[1], b3[1], b4[1], b5[1])
  af <- 1 - exp(-bvarcen %*% coef_vec)
  af[af < 0] <- 0  # Gasparrini & Leone (2014): range-parameter approach
  daily_deaths <- annual_deaths[1] / n_days_year[1]
  list(as.vector(af), as.vector(af * daily_deaths))
}, by = .(age_group, year)]

proj_long[, temp_range := classify_temp_range(temp, p025_hist, p975_hist, mmt)]

cat("=== Daily RR/AF Computed (Masselot-style) ===\n")
print(head(proj_long[, .(date, year, age_group, temp, mmt, af_day, an_day, temp_range)], 10))
```

---

# Aggregate to Annual ANs with CIs

## Point Estimates by Age Group and Temperature Range

```{r annual-ans-point}
annual_ans <- proj_long[, .(
  an_total = sum(an_day, na.rm = TRUE),
  annual_deaths = unique(annual_deaths)),
  by = .(year, age_group, temp_range)]

annual_ans[, af := an_total / annual_deaths]

cat("=== Annual AN Point Estimates ===\n")
print(head(annual_ans, 20))
```

## Confidence Intervals via Pre-Simulated Coefficients

Following Masselot et al., we use the 1000 pre-simulated coefficient sets (`coef_simu.csv`) from the second-stage meta-regression to compute empirical 95% CIs.

```{r compute-ci}
proj_long[, daily_deaths := annual_deaths / n_days_year]
temp_ranges <- c("extreme_cold", "moderate_cold", "moderate_heat", "extreme_heat")

# For each year x age_group, compute 1000 simulated ANs
ci_list <- list()
years <- unique(proj_long$year)

for (yr in years) {
  yr_data <- proj_long[year == yr]
  age_groups_u <- unique(yr_data$age_group)

  for (ag in age_groups_u) {
    ag_data <- yr_data[age_group == ag]

    # Build centred basis once for this age group's full-year temperatures
    bvar <- do.call(onebasis, c(list(x = ag_data$temp), argvar))
    cenvec <- do.call(onebasis,
      c(list(x = ag_data$mmt[1]), argvar))
    bvarcen <- scale(bvar, center = cenvec, scale = FALSE)

    # Pre-compute range indicators
    range_idx <- list(
      extreme_cold = which(ag_data$temp_range == "extreme_cold"),
      moderate_cold = which(ag_data$temp_range == "moderate_cold"),
      moderate_heat = which(ag_data$temp_range == "moderate_heat"),
      extreme_heat = which(ag_data$temp_range == "extreme_heat"))

    # Get 1000 simulated coefficient vectors for this age group
    sim_coefs <- coef_simu_city[agegroup == ag]
    n_sim <- uniqueN(sim_coefs$sim)
    dd <- ag_data$daily_deaths[1]

    # Compute AN for each simulation
    an_mat <- t(sapply(seq_len(n_sim), function(s) {
      cf <- as.numeric(sim_coefs[sim == s, .(b1, b2, b3, b4, b5)])
      af_sim <- 1 - exp(-bvarcen %*% cf)
      an_sim <- as.vector(af_sim * dd)
      c(total = sum(an_sim),
        extreme_cold = sum(an_sim[range_idx$extreme_cold]),
        moderate_cold = sum(an_sim[range_idx$moderate_cold]),
        moderate_heat = sum(an_sim[range_idx$moderate_heat]),
        extreme_heat = sum(an_sim[range_idx$extreme_heat]))
    }))

    # Quantiles: extract low and hi into separate columns
    ci_vals <- apply(an_mat, 2, quantile, c(.025, .975), na.rm = TRUE)
    # ci_vals is 2 x 5 matrix: rows = low/hi, cols = total/extreme_cold/...
    ci_row <- data.table(
      year = yr, age_group = ag,
      total_low = ci_vals["2.5%", "total"],
      total_hi = ci_vals["97.5%", "total"],
      extreme_cold_low = ci_vals["2.5%", "extreme_cold"],
      extreme_cold_hi = ci_vals["97.5%", "extreme_cold"],
      moderate_cold_low = ci_vals["2.5%", "moderate_cold"],
      moderate_cold_hi = ci_vals["97.5%", "moderate_cold"],
      moderate_heat_low = ci_vals["2.5%", "moderate_heat"],
      moderate_heat_hi = ci_vals["97.5%", "moderate_heat"],
      extreme_heat_low = ci_vals["2.5%", "extreme_heat"],
      extreme_heat_hi = ci_vals["97.5%", "extreme_heat"])
    ci_list[[length(ci_list) + 1L]] <- ci_row
  }
}

ci_dt <- rbindlist(ci_list)

# Reshape to long: one row per year x age_group x temp_range
ci_long <- melt(ci_dt, id.vars = c("year", "age_group"),
  variable.name = "measure", value.name = "value")
ci_long[, temp_range := gsub("_(low|hi)$", "", measure)]
ci_long[, bound := gsub("^.*_(low|hi)$", "\\1", measure)]
ci_long[, measure := NULL]
ci_wide <- dcast(ci_long, year + age_group + temp_range ~ bound,
  value.var = "value")

cat("=== CI Computation Complete ===\n")
cat(sprintf("  CI rows: %s\n", format(nrow(ci_wide), big.mark = ",")))
print(head(ci_wide, 10))
```

## Merge Point Estimates with CIs

```{r merge-ci}
annual_ans <- merge(annual_ans, ci_wide, by = c("year", "age_group", "temp_range"))

# Add metadata columns
annual_ans[, `:=`(
  city = city,
  city_code = city_code,
  gcm = demo_gcm,
  ssp = scenario_label,
  adaptation = adaptation_level)]

# Rename to Masselot-style output convention
setnames(annual_ans,
  old = c("an_total", "low", "hi"),
  new = c("an_est", "an_low", "an_hi"))

setcolorder(annual_ans, c(
  "city", "city_code", "gcm", "ssp", "adaptation",
  "year", "age_group", "temp_range", "annual_deaths",
  "an_est", "an_low", "an_hi", "af"))

cat("=== Final Annual AN Table (Masselot-style output) ===\n")
print(head(annual_ans, 20))
```

## Annual Totals Across Temperature Ranges

```{r annual-totals}
annual_totals <- annual_ans[, .(
  annual_deaths = unique(annual_deaths),
  an_est = sum(an_est),
  an_low = sum(an_low),
  an_hi = sum(an_hi),
  af = sum(an_est) / unique(annual_deaths)),
  by = .(year, age_group)]

cat("=== Annual Totals ===\n")
print(head(annual_totals, 20))
```

---

# Diagnostics and Visualisation

## Annual Attributable Numbers by Temperature Range

```{r plot-annual-ans-range}
ggplot(annual_ans, aes(x = year, y = an_est, color = temp_range)) +
  geom_ribbon(aes(ymin = an_low, ymax = an_hi, fill = temp_range),
    alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  facet_wrap(~ age_group, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(
    title = sprintf("Projected Annual Attributable Numbers — %s", city),
    subtitle = sprintf("%s | %s | No adaptation | Wide age groups", scenario_label, demo_gcm),
    x = "Year", y = "Attributable deaths (AN)",
    color = "Temperature range", fill = "Temperature range") +
  theme_minimal(base_size = 12)
```

## Age Pattern in a Selected Year

```{r age-pattern-single-year}
selected_year <- min(annual_ans$year)
age_plot_dt <- annual_ans[year == selected_year]

ggplot(age_plot_dt, aes(x = age_group, y = an_est, fill = temp_range)) +
  geom_col(position = "stack") +
  scale_y_continuous(labels = comma) +
  labs(
    title = sprintf("AN by Age Group — %d", selected_year),
    subtitle = sprintf("%s | %s | %s", city, scenario_label, demo_gcm),
    x = "Age group", y = "Attributable deaths (AN)",
    fill = "Temperature range") +
  theme_minimal(base_size = 12)
```

---

# Export Results

```{r export-results}
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

fwrite(annual_ans, output_file_annual)

# Daily-level output (point estimates only, no CI at daily level)
fwrite(proj_long[, .(
  city = city, city_code = city_code,
  gcm = demo_gcm, ssp = scenario_label, adaptation = adaptation_level,
  date, year, doy, age_group, temp, mmt, af_day, an_day, temp_range)],
  output_file_daily)

cat("=== Files Written ===\n")
cat(sprintf("  Annual AN table (with CIs): %s\n", output_file_annual))
cat(sprintf("  Daily attribution table: %s\n", output_file_daily))
```

---

# Notebook Summary

This notebook has:

1. loaded city-specific wide-age-group ERFs for Madrid,
2. reconstructed temperature-mortality RR curves using B-spline basis (knots at 10th, 75th, 90th percentiles),
3. derived age-group-specific MMTs,
4. classified projected daily temperatures into four standard ranges,
5. computed daily RR, AF, and attributable numbers following the Masselot-style `1 - exp(-bvarcen %*% coef)` approach,
6. computed empirical 95% CIs using 1000 pre-simulated coefficient sets from the second-stage meta-regression,
7. aggregated results into annual ANs and AFs by age group and temperature range.

## Main deliverable for the next notebook

The main output is the annual AN table by:

- year
- age group
- temperature range

with point estimates and 95% confidence intervals.

The next notebook in the series will take this output and **disaggregate the age-group ANs into single years of age using `pclm`**.

---

# Session Information

```{r session-info}
sessionInfo()
```