---
title: "Disaggregate Annual ANs to Single Years of Age Using PCLM"
subtitle: "Part 2: Age-Group to Single-Age Disaggregation"
author: "Daniel Precioso"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    fig_width: 10
    fig_height: 6
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    keep_tex: false
---

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = FALSE,
  comment = "##",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook takes the annual attributable numbers (ANs) by wide age group from Notebook 1 and **disaggregates them to single years of age** using the **Penalized Composite Link Model (PCLM)**.

The PCLM (Rizzi et al. 2015) estimates a smooth latent distribution from coarsely grouped count data. It treats the observed age-group ANs as Poisson-distributed aggregates of an underlying smooth single-age sequence, and recovers that sequence through penalized maximum likelihood.

## Input

- `madrid_projected_ans_annual.csv` — annual ANs by year, age group (20–44, 45–64, 65–74, 75–84, 85+), and temperature range (extreme_cold, moderate_cold, moderate_heat, extreme_heat)

## Output

- `madrid_projected_ans_single_age.csv` — annual ANs by year, single age (20–100), and temperature range

---

# 1. Setup and Configuration

```{r setup-libraries}
library(data.table)
library(ungroup)
library(ggplot2)
library(scales)

cat("Libraries loaded successfully.\n")
```

```{r set-config}
# Demo configuration (must match Notebook 1)
city <- "Madrid"
city_code <- "ES001C"

# Shared output directory (same across all notebooks)
output_dir <- "../results/demo"
city_prefix <- tolower(gsub(" ", "_", city))

# Input paths
input_annual_ans <- file.path(output_dir, paste0(city_prefix, "_ans_annual.csv"))

# Output paths
output_single_age <- file.path(output_dir, paste0(city_prefix, "_ans_single_age.csv"))
output_diagnostic <- file.path(output_dir, paste0(city_prefix, "_pclm_diagnostics.csv"))

# PCLM parameters
pclm_nlast <- 16               # Last open interval 85+ capped at 100 (ages 85..100 = 16 years)
pclm_lambda <- NA              # Auto-select smoothing parameter via BIC
pclm_deg <- 3L                 # B-spline degree
pclm_kr <- 2L                  # Knot ratio (intervals per knot)

# Age group boundaries (start of each interval)
age_group_starts <- c(20, 45, 65, 75, 85)
age_group_labels <- c("20-44", "45-64", "65-74", "75-84", "85+")
single_ages <- 20:100

cat("=== Configuration ===\n")
cat(sprintf("  City: %s (%s)\n", city, city_code))
cat(sprintf("  Output directory: %s\n", output_dir))
cat(sprintf("  Age groups: %s\n", paste(age_group_labels, collapse = ", ")))
cat(sprintf("  Single ages: %d–%d (%d ages)\n",
            min(single_ages), max(single_ages), length(single_ages)))
cat(sprintf("  PCLM knot ratio: %d, degree: %d\n", pclm_kr, pclm_deg))
```

---

# 2. Load Annual ANs from Notebook 1

```{r load-annual-ans}
annual_ans <- fread(input_annual_ans)

cat("=== Annual ANs Loaded ===\n")
cat(sprintf("  Rows: %s\n", format(nrow(annual_ans), big.mark = ",")))
cat(sprintf("  Years: %d–%d\n", min(annual_ans$year), max(annual_ans$year)))
cat(sprintf("  Age groups: %s\n", paste(unique(annual_ans$age_group), collapse = ", ")))
cat(sprintf("  Temp ranges: %s\n", paste(unique(annual_ans$temp_range), collapse = ", ")))
```

---

# 3. PCLM Disaggregation

For each combination of year and temperature range, we extract the five age-group ANs and apply the PCLM to recover the single-age distribution.

## 3.1 Define the PCLM Wrapper

```{r pclm-wrapper}
disaggregate_an <- function(an_values, age_starts = age_group_starts,
                            nlast = pclm_nlast, lambda = pclm_lambda,
                            deg = pclm_deg, kr = pclm_kr) {
  M <- pclm(
    x = age_starts,
    y = an_values,
    nlast = nlast,
    out.step = 1,
    control = list(
      lambda = lambda,
      kr = kr,
      deg = deg,
      max.iter = 200,
      tol = 1e-8,
      opt.method = "BIC"
    )
  )
  as.numeric(fitted(M))
}
```

## 3.2 Run Disaggregation for All Year–Temp-Range Combinations

```{r run-pclm-all}
results_list <- list()
diagnostics_list <- list()

combos <- unique(annual_ans[, .(year, temp_range)])
n_combos <- nrow(combos)

for (i in seq_len(n_combos)) {
  yr <- combos$year[i]
  tr <- combos$temp_range[i]

  subset_dt <- annual_ans[year == yr & temp_range == tr]
  an_vec <- subset_dt$AN

  if (length(an_vec) != 5L) {
    warning(sprintf("Skipping year=%d, temp_range=%s: %d age groups (expected 5)",
                    yr, tr, length(an_vec)))
    next
  }

  single_an <- disaggregate_an(an_vec)

  results_list[[length(results_list) + 1L]] <- data.table(
    year = yr,
    age = single_ages,
    temp_range = tr,
    AN = single_an
  )

  # Store diagnostics
  diagnostics_list[[length(diagnostics_list) + 1L]] <- data.table(
    year = yr,
    temp_range = tr,
    total_an_input = sum(an_vec),
    total_an_output = sum(single_an),
    n_age_groups = 5L,
    n_single_ages = length(single_an)
  )
}

single_age_an <- rbindlist(results_list)
diagnostics <- rbindlist(diagnostics_list)

cat("=== PCLM Disaggregation Complete ===\n")
cat(sprintf("  Combinations processed: %d\n", n_combos))
cat(sprintf("  Single-age AN rows: %s\n", format(nrow(single_age_an), big.mark = ",")))
```

## 3.3 Verify Conservation of Total AN

```{r verify-totals}
diagnostics[, diff := total_an_input - total_an_output]
diagnostics[, relative_diff := diff / total_an_input * 100]

cat("=== Conservation Check ===\n")
cat(sprintf("  Max absolute difference: %.6f\n", max(abs(diagnostics$diff))))
cat(sprintf("  Max relative diff (pct): %.10f\n", max(abs(diagnostics$relative_diff))))
cat(sprintf("  All relative diffs < 1e-6: %s\n", all(abs(diagnostics$diff) < 1e-6)))

# Verify that within each year, the original age-group totals are approximately preserved
# by re-aggregating the single-age ANs back into the wide age groups
temp_check <- single_age_an[year == 2020 & temp_range == "moderate_heat"]
reagg <- data.table(
  age_group = age_group_labels,
  original = annual_ans[year == 2020 & temp_range == "moderate_heat"]$AN,
  reconstructed = c(
    sum(temp_check[age >= 20 & age <= 44]$AN),
    sum(temp_check[age >= 45 & age <= 64]$AN),
    sum(temp_check[age >= 65 & age <= 74]$AN),
    sum(temp_check[age >= 75 & age <= 84]$AN),
    sum(temp_check[age >= 85 & age <= 100]$AN)
  )
)
reagg[, diff := original - reconstructed]

cat("\n=== Re-aggregation Verification (2020 moderate_heat) ===\n")
print(reagg)
```

---

# 4. Add Metadata Columns

```{r add-metadata}
meta <- unique(annual_ans[, .(city, city_code, gcm, ssp, adaptation)])
for (col in names(meta)) {
  set(single_age_an, j = col, value = meta[[col]][1L])
}
setcolorder(single_age_an, c("city", "city_code", "gcm", "ssp", "adaptation",
                              "year", "age", "temp_range", "AN"))

cat("=== Single-Age AN Table ===\n")
print(head(single_age_an, 10))
```

---

# 5. Diagnostics and Visualisation

## 5.1 Compare Age-Group vs Single-Age Distributions

```{r plot-age-group-vs-single}
plot_years <- c(2020, 2050, 2099)

plot_data <- single_age_an[year %in% plot_years]

ggplot(plot_data, aes(x = age, y = AN, color = factor(year))) +
  geom_line(linewidth = 1) +
  facet_wrap(~ temp_range, scales = "free_y") +
  labs(
    title = "Single-Age Attributable Numbers (PCLM Disaggregation)",
    subtitle = sprintf("%s | %s | %s",
                       unique(plot_data$city),
                       unique(plot_data$ssp),
                       unique(plot_data$gcm)),
    x = "Age",
    y = "Attributable number (AN)",
    color = "Year"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal(base_size = 12)
```

## 5.2 Heatmap of Single-Age ANs Over Time

```{r plot-heatmap}
# Pivot to wide for heatmap: year x age for moderate_heat (most impactful)
heat_data <- single_age_an[temp_range %in% c("extreme_heat", "extreme_cold")]

ggplot(heat_data, aes(x = year, y = age, fill = AN)) +
  geom_tile() +
  facet_wrap(~ temp_range, ncol = 1) +
  scale_fill_gradient2(low = "white", high = "red", labels = comma) +
  labs(
    title = "Single-Age ANs Over Time by Temperature Range",
    subtitle = sprintf("%s | %s | %s",
                       unique(heat_data$city),
                       unique(heat_data$ssp),
                       unique(heat_data$gcm)),
    x = "Year",
    y = "Age",
    fill = "AN"
  ) +
  theme_minimal(base_size = 12)
```

## 5.3 Cross-Section: Age Profile in Selected Years

```{r plot-cross-section}
cross_data <- single_age_an[year %in% plot_years &
                            temp_range %in% c("moderate_heat", "extreme_heat", "moderate_cold")]

ggplot(cross_data, aes(x = age, y = AN, color = factor(year), linetype = temp_range)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Age Profile of Attributable Mortality",
    subtitle = sprintf("%s | %s", unique(cross_data$city), unique(cross_data$ssp)),
    x = "Age",
    y = "Attributable number (AN)",
    color = "Year",
    linetype = "Temperature range"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal(base_size = 12)
```

---

# 6. Export Results

```{r export-results}
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

fwrite(single_age_an, output_single_age)
fwrite(diagnostics, output_diagnostic)

cat("=== Files Written ===\n")
cat(sprintf("  Single-age AN table: %s\n", output_single_age))
cat(sprintf("  Diagnostics: %s\n", output_diagnostic))
```

---

# 7. Notebook Summary

This notebook:

1. loaded the annual attributable numbers (ANs) by wide age group from Notebook 1,
2. applied the Penalized Composite Link Model (PCLM) to disaggregate each year–temperature-range combination from 5 wide age groups (20–44, 45–64, 65–74, 75–84, 85+) to 81 single ages (20–100),
3. verified that the total AN is conserved within numerical tolerance,
4. produced diagnostic plots showing the smooth age distributions.

## Main deliverable for the next notebook

The file `madrid_projected_ans_single_age.csv` contains annual ANs by year, single age, and temperature range. The next notebook in the series will combine these with all-cause mortality and population data to construct the "dataset for analysis" and derive period life-table inputs.

---

# Session Information

```{r session-info}
sessionInfo()
```