---
title: "Construct the Dataset for Analysis and Derive Period Life-Table Inputs"
subtitle: "Part 3: All-Cause Mortality, Population, and Temperature-Attributable Mortality"
author: "Daniel Precioso"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = FALSE,
  comment = "##",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook combines three data streams into a single **analysis dataset**:

1. **Single-age attributable numbers (ANs)** by temperature range (from Notebook 2)
2. **Single-age all-cause mortality** for Madrid
3. **Single-age population** for Madrid

The dataset is structured following the approach described in the life-table decomposition literature (Lloyd-style supplementary table), with separate columns for all-cause mortality and temperature-attributable mortality by range.

## Inputs

- `results/demo/madrid_ans_single_age.csv` — disaggregated ANs
- `data/city_results.csv` — baseline Madrid population and deaths by age group
- `data/ES_mortality_male.csv` — national single-age mortality trends
- `data/coefs.csv` — ERF coefficients (for reference; ANs are pre-computed)

## Outputs

- `results/demo/madrid_analysis_dataset.csv` — combined analysis dataset
- `results/demo/madrid_lifetable_inputs.csv` — mx and population by year

---

# Setup and Configuration

```{r setup-libraries}
library(data.table)
library(ungroup)
library(ggplot2)
library(scales)
library(eurostat)

cat("Libraries loaded successfully.\n")
```

```{r set-config}
# Demo configuration (must match Notebook 1)
city <- "Madrid"
city_code <- "ES001C"
sex <- "M"                         # "M", "F", or "T"
ssp_label <- "SSP3-7.0"
gcm <- NA_character_  # will be read from input data

# Detect country code from city metadata
city_meta <- fread("../data/city_results.csv")
country_code <- unique(city_meta[URAU_CODE == city_code, CNTR_CODE])[1]

# Shared output directory (same across all notebooks)
output_dir <- "../results/demo"
city_prefix <- tolower(gsub(" ", "_", city))

# Input paths
input_single_age_an <- file.path(output_dir, paste0(city_prefix, "_ans_single_age.csv"))
input_city_results <- "../data/city_results.csv"
input_coefs <- "../data/coefs.csv"

# Output paths
output_dataset <- file.path(output_dir, paste0(city_prefix, "_analysis_dataset.csv"))
output_lifetable_inputs <- file.path(output_dir, paste0(city_prefix, "_lifetable_inputs.csv"))
output_pop_single <- file.path(output_dir, paste0(city_prefix, "_population_single_age.csv"))

# Age configuration
age_groups <- c("20-44", "45-64", "65-74", "75-84", "85+")
age_group_starts <- c(20, 45, 65, 75, 85)
single_ages <- 20:100
n_ages <- length(single_ages)
pclm_nlast <- 16  # ages 85-100

cat("=== Configuration ===\n")
cat(sprintf("  City: %s (%s)\n", city, city_code))
cat(sprintf("  Country: %s | Sex: %s\n", country_code, sex))
cat(sprintf("  Output directory: %s\n", output_dir))
cat(sprintf("  Single ages: %d–%d\n", min(single_ages), max(single_ages)))
```

---

# Load Pre-Computed Single-Age ANs

```{r load-single-age-ans}
single_an <- fread(input_single_age_an)

# Set a key for merging
setkey(single_an, year, age, temp_range)

cat("=== Single-Age ANs Loaded ===\n")
cat(sprintf("  Rows: %s\n", format(nrow(single_an), big.mark = ",")))
cat(sprintf("  Years: %d–%d\n", min(single_an$year), max(single_an$year)))
cat(sprintf("  Ages: %d–%d\n", min(single_an$age), max(single_an$age)))

# Read GCM, SSP, adaptation from the data
gcm <- unique(single_an$gcm)[1]
ssp_label <- unique(single_an$ssp)[1]

cat(sprintf("  GCM: %s | SSP: %s\n", gcm, ssp_label))
```

---

# Build Single-Age Population for Madrid

We use the PCLM to disaggregate the five wide-age-group populations from `city_results.csv` into a smooth single-age distribution. The disaggregated population is held at the baseline age structure and scaled by the total population trend from `parsed_pop_all.csv`.

```{r build-population}
city_res <- fread(input_city_results)
madrid_base <- city_res[URAU_CODE == city_code & agegroup %in% age_groups]

base_pop <- setNames(madrid_base$agepop, madrid_base$agegroup)
base_death <- setNames(madrid_base$death, madrid_base$agegroup)
base_mx <- base_death / base_pop  # age-group mx

# Disaggregate population to single ages using PCLM
pop_pclm <- pclm(x = age_group_starts, y = base_pop, nlast = pclm_nlast,
                 out.step = 1, control = list(lambda = NA, kr = 2, deg = 3))
pop_single_age <- as.numeric(fitted(pop_pclm))
pop_single_age <- pmax(pop_single_age, 0)
pop_single_age <- pop_single_age / sum(pop_single_age) * sum(base_pop)  # renormalise

# Disaggregate deaths to single ages using PCLM
death_pclm <- pclm(x = age_group_starts, y = base_death, nlast = pclm_nlast,
                   out.step = 1, control = list(lambda = NA, kr = 2, deg = 3))
death_single_age <- as.numeric(fitted(death_pclm))
death_single_age <- pmax(death_single_age, 0)
death_single_age <- death_single_age / sum(death_single_age) * sum(base_death)

# Baseline single-age mx
mx_single_age_baseline <- death_single_age / pop_single_age
mx_single_age_baseline[is.nan(mx_single_age_baseline) | is.infinite(mx_single_age_baseline)] <- 0

pop_dt <- data.table(age = single_ages, population = round(pop_single_age),
                     mx_baseline = mx_single_age_baseline)

cat("=== Single-Age Population Built ===\n")
print(head(pop_dt, 10))
cat(sprintf("  Total population (20+): %.0f\n", sum(pop_dt$population)))
```

---

# Build Projected All-Cause Mortality

We project Madrid's single-age all-cause mortality using the **Eurostat EUROPOP2019** projections via the `load_eurostat_mortality()` function. This downloads:
- `demo_mlifetable` — historical age-specific death rates (DEATHRATE) for baseline mx
- `proj_19np` — population projections for future mortality trend
- Returns projected mx for any country and sex, ages 20–100, years 2020–2100

```{r build-mortality}
source("../R/load_data.R")

proj_mx <- load_eurostat_mortality(country_code, sex = sex,
  age_min = min(single_ages), age_max = max(single_ages),
  year_min = min(single_an$year), year_max = max(single_an$year))

cat("=== Projected All-Cause Mortality (Eurostat) ===\n")
cat(sprintf("  Country: %s | Sex: %s | Years: %d–%d\n",
  country_code, sex, min(proj_mx$year), max(proj_mx$year)))
cat(sprintf("  Ages: %d–%d | Rows: %s\n",
  min(proj_mx$age), max(proj_mx$age), format(nrow(proj_mx), big.mark = ",")))
print(proj_mx[year %in% c(2022, 2050, 2099) & age %in% c(20, 50, 80)])
```

---

# Build Analysis Dataset

Combine population, all-cause mortality, and single-age ANs into the final analysis table.

```{r build-dataset}
# Start with population and all-cause mx
# Combine Madrid baseline mx with Eurostat mortality trend
proj_mx <- merge(proj_mx[, .(age, year, mx_ratio)],
  pop_dt[, .(age, mx_baseline)], by = "age")
proj_mx[, mx := mx_baseline * mx_ratio]

analysis_dt <- proj_mx[, .(year, age, mx, population = pop_dt$population[match(age, pop_dt$age)])]

# Add annual deaths
analysis_dt[, deaths := mx * population]

# Melt single-age ANs from wide temp_range to columns
an_wide <- dcast(single_an, year + age ~ temp_range, value.var = "AN", fill = 0)
setnames(an_wide,
         old = c("extreme_cold", "moderate_cold", "moderate_heat", "extreme_heat"),
         new = c("an_extreme_cold", "an_moderate_cold", "an_moderate_heat", "an_extreme_heat"))

# Merge into analysis dataset
analysis_dt <- merge(analysis_dt, an_wide, by = c("year", "age"), all.x = TRUE)

# Fill missing ANs with 0
for (col in c("an_extreme_cold", "an_moderate_cold", "an_moderate_heat", "an_extreme_heat")) {
  if (col %in% names(analysis_dt)) {
    analysis_dt[is.na(get(col)), (col) := 0]
  }
}

# Compute total attributable deaths
analysis_dt[, an_total := an_extreme_cold + an_moderate_cold + an_moderate_heat + an_extreme_heat]

# Compute baseline (non-attributable) deaths
analysis_dt[, deaths_baseline := deaths - an_total]
analysis_dt[deaths_baseline < 0, deaths_baseline := 0]

# Add metadata
analysis_dt[, `:=`(
  city = city,
  city_code = city_code,
  gcm = gcm,
  ssp = ssp_label
)]

setcolorder(analysis_dt, c("city", "city_code", "gcm", "ssp",
                            "year", "age", "population",
                            "deaths", "mx",
                            "deaths_baseline",
                            "an_total",
                            "an_extreme_cold", "an_moderate_cold",
                            "an_moderate_heat", "an_extreme_heat"))
setorder(analysis_dt, year, age)

cat("=== Analysis Dataset Built ===\n")
cat(sprintf("  Rows: %s\n", format(nrow(analysis_dt), big.mark = ",")))
cat(sprintf("  Years: %d–%d\n", min(analysis_dt$year), max(analysis_dt$year)))
cat(sprintf("  Ages: %d–%d\n", min(analysis_dt$age), max(analysis_dt$age)))
print(head(analysis_dt[year == 2022 & age %in% c(20, 30, 40, 50, 60, 70, 80)], 10))
```

---

# Derive Period Life-Table Inputs

For each year, compute period life tables (baseline and climate-adjusted) using the standard demographic approach.

```{r derive-lifetable-inputs}
source("../R/period_lifetable.R")
source("../R/utils.R")

lifetable_years <- sort(unique(analysis_dt$year))

lt_list_base <- list()
lt_list_clim <- list()

for (yr in lifetable_years) {
  yr_data <- analysis_dt[year == yr]
  
  # Baseline life table (no temperature-attributable mortality)
  mx_base <- yr_data$deaths_baseline / yr_data$population
  mx_base[is.nan(mx_base) | is.infinite(mx_base)] <- 0
  lt_base <- lifetable(mx = mx_base, age = yr_data$age, ax = 0.5)
  lt_base[, year := yr]
  lt_list_base[[length(lt_list_base) + 1L]] <- lt_base
  
  # Climate-adjusted life table (with temperature-attributable mortality)
  mx_clim <- yr_data$deaths / yr_data$population
  mx_clim[is.nan(mx_clim) | is.infinite(mx_clim)] <- 0
  lt_clim <- lifetable(mx = mx_clim, age = yr_data$age, ax = 0.5)
  lt_clim[, year := yr]
  lt_list_clim[[length(lt_list_clim) + 1L]] <- lt_clim
}

lt_base_all <- rbindlist(lt_list_base)
lt_clim_all <- rbindlist(lt_list_clim)

# Extract life expectancy at age 65 (following Lloyd et al. 2024)
le_base <- lt_base_all[age == 65, .(year, le_base = ex)]
le_clim <- lt_clim_all[age == 65, .(year, le_clim = ex)]

cat("=== Period Life Tables Computed ===\n")
cat(sprintf("  Years: %d–%d\n", min(lifetable_years), max(lifetable_years)))
print(merge(le_base, le_clim, by = "year")[year %in% c(2022, 2050, 2099)])
```

---

# Diagnostics

## Life Expectancy at Age 65 Over Time

```{r plot-life-expectancy}
le_all <- merge(le_base, le_clim, by = "year")

ggplot(melt(le_all, id.vars = "year", variable.name = "scenario", value.name = "ex"),
       aes(x = year, y = ex, color = scenario)) +
  geom_line(linewidth = 1) +
  labs(
    title = sprintf("Period Life Expectancy at Age 65 — %s", city),
    subtitle = sprintf("%s | %s | GCM: %s", city_code, ssp_label, gcm),
    x = "Year",
    y = "Life expectancy at age 20",
    color = "Scenario"
  ) +
  scale_color_manual(
    values = c("le_base" = "steelblue", "le_clim" = "tomato"),
    labels = c("Baseline (no temperature)", "Climate-adjusted")
  ) +
  theme_minimal(base_size = 12)
```

---

# Export Results

```{r export}
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

fwrite(analysis_dt, output_dataset)
fwrite(pop_dt, output_pop_single)

# Life-table inputs: mx by year and age
lt_inputs <- merge(
  analysis_dt[, .(year, age, population, mx, mx_baseline = deaths_baseline / population)],
  lt_base_all[, .(year, age, qx_base = qx, lx_base = lx, ex_base = ex)],
  by = c("year", "age"), all.x = TRUE
)
lt_inputs <- merge(
  lt_inputs,
  lt_clim_all[, .(year, age, qx_clim = qx, lx_clim = lx, ex_clim = ex)],
  by = c("year", "age"), all.x = TRUE
)
setorder(lt_inputs, year, age)
fwrite(lt_inputs, output_lifetable_inputs)

cat("=== Files Written ===\n")
cat(sprintf("  Analysis dataset: %s\n", output_dataset))
cat(sprintf("  Period life-table inputs: %s\n", output_lifetable_inputs))
cat(sprintf("  Single-age population: %s\n", output_pop_single))
```

---

# Notebook Summary

This notebook:

1. loaded single-age attributable numbers (ANs) from Notebook 2,
2. built single-age population for Madrid using PCLM disaggregation of the city_results.csv age groups,
3. built projected all-cause mortality by applying national mortality trends from `ES_mortality_male.csv` to Madrid's baseline,
4. combined all streams into a single analysis dataset with all-cause mortality, baseline mortality, and temperature-attributable mortality by range,
5. derived period life tables (baseline and climate-adjusted) for each projection year,
6. computed life expectancy at age 20 for both scenarios.

## Main deliverable for the next notebook

The analysis dataset (`madrid_analysis_dataset.csv`) and period life-table inputs (`madrid_period_lifetable_inputs.csv`) are ready. Notebook 4 will apply the Lloyd-style decomposition to attribute changes in life expectancy and lifespan inequality to age groups and temperature ranges.

---

# Session Information

```{r session-info}
sessionInfo()
```