---
title: "Decompose Changes in Life Expectancy and Lifespan Inequality"
subtitle: "Part 4: Lloyd-Style Decomposition by Age and Temperature Range"
author: "Daniel Precioso"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = FALSE,
  comment = "##",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook applies a **Lloyd/Aburto-style decomposition** using `DemoDecomp::horiuchi()` to separate the projected impact of temperature-attributable mortality on **life expectancy (LE) at age 65** and **lifespan inequality (LI) among people aged 65+** into contributions by:

- **single years of age** (65--100)
- **temperature range** (extreme cold, moderate cold, moderate heat, extreme heat) and "rest" (non-temperature deaths)

The decomposition uses the continuous stepwise-replacement (Horiuchi) method with 50 integration steps, matching Lloyd et al. (2024) and Aburto et al. The cause-specific mortality rates are passed directly to `horiuchi()`, which simultaneously attributes the change across both age and cause dimensions.

## Input

- `results/demo/madrid_analysis_dataset.csv` — analysis dataset with population, mx, and ANs

## Outputs

- `results/demo/madrid_le_decomposition.csv` — LE decomposition by age and temp range
- `results/demo/madrid_li_decomposition.csv` — LI decomposition by age and temp range
- `results/demo/madrid_lifespan_inequality.csv` — LI measures over time

---

# Setup and Configuration

```{r setup-libraries}
library(data.table)
library(ggplot2)
library(scales)
library(DemoDecomp)

source("../R/period_lifetable.R")
source("../R/utils.R")

cat("Libraries loaded successfully.\n")
```

```{r set-config}
city <- "Madrid"
city_code <- "ES001C"

# Focus on ages 65+ (following Lloyd et al. 2024)
cond_age <- 65L
cond_ages <- cond_age:100

output_dir <- "../results/demo"
city_prefix <- tolower(gsub(" ", "_", city))

input_dataset <- file.path(output_dir, paste0(city_prefix, "_analysis_dataset.csv"))

output_le_decomp <- file.path(output_dir, paste0(city_prefix, "_le_decomposition.csv"))
output_li_decomp <- file.path(output_dir, paste0(city_prefix, "_li_decomposition.csv"))
output_li_series <- file.path(output_dir, paste0(city_prefix, "_lifespan_inequality.csv"))

decomp_years <- c(2022, 2050, 2099)

temp_ranges <- c("an_extreme_cold", "an_moderate_cold",
  "an_moderate_heat", "an_extreme_heat")
temp_labels <- c("extreme_cold", "moderate_cold",
  "moderate_heat", "extreme_heat")

cat("=== Configuration ===\n")
cat(sprintf("  City: %s (%s)\n", city, city_code))
cat(sprintf("  Conditional age: %d+\n", cond_age))
cat(sprintf("  Decomposition years: %s\n", paste(decomp_years, collapse = ", ")))
```

---

# Load Analysis Dataset

```{r load-data}
analysis <- fread(input_dataset)
analysis <- analysis[order(year, age)]

cat("=== Analysis Dataset Loaded ===\n")
cat(sprintf("  Rows: %s\n", format(nrow(analysis), big.mark = ",")))
cat(sprintf("  Years: %d–%d\n", min(analysis$year), max(analysis$year)))
cat(sprintf("  Ages: %d–%d\n", min(analysis$age), max(analysis$age)))
```

---

# Core Life-Table Functions (Lloyd-style, piecewise-exponential)

These functions are restricted to ages 65+ as recommended by Lloyd et al. (2024). Remaining life expectancy at age 65, $e_{65}$, and the standard deviation of age at death among people aged 65+ are computed from a piecewise-constant hazard life table.

```{r core-functions}
le65_from_mx <- function(mx_vec) {
  ages <- cond_ages
  nx <- c(diff(ages), Inf)
  px <- exp(-mx_vec * nx)
  lx <- cumprod(c(1, head(px, -1)))
  dx <- c(-diff(lx), tail(lx, 1))
  Lx <- fifelse(mx_vec == 0, lx * nx, dx / mx_vec)
  Tx <- rev(cumsum(rev(Lx)))
  ex <- Tx / lx
  ex[1]
}

sd65_from_mx <- function(mx_vec) {
  ages <- cond_ages
  nx <- c(diff(ages), Inf)
  px <- exp(-mx_vec * nx)
  lx <- cumprod(c(1, head(px, -1)))
  dx <- c(-diff(lx), tail(lx, 1))
  ex65 <- le65_from_mx(mx_vec)
  sqrt(sum(dx * (ages + 0.5 - ex65)^2) / lx[1])
}

le65_cod_fun <- function(mx_flat, ages, n_causes) {
  dim(mx_flat) <- c(length(ages), n_causes)
  mx_all <- rowSums(mx_flat)
  le65_from_mx(mx_all)
}

sd65_cod_fun <- function(mx_flat, ages, n_causes) {
  dim(mx_flat) <- c(length(ages), n_causes)
  mx_all <- rowSums(mx_flat)
  sd65_from_mx(mx_all)
}
```

---

# Compute Baseline and Climate-Adjusted LE and LI

```{r compute-le-li}
all_years <- sort(unique(analysis$year))

stats_list <- list()
for (yr in all_years) {
  yr_data <- analysis[year == yr & age >= cond_age]

  mx_base <- yr_data$deaths_baseline / yr_data$population
  mx_base[is.nan(mx_base) | is.infinite(mx_base)] <- 1e-10

  mx_clim <- yr_data$deaths / yr_data$population
  mx_clim[is.nan(mx_clim) | is.infinite(mx_clim)] <- 1e-10

  stats_list[[length(stats_list) + 1L]] <- data.table(
    year = yr,
    e65_base = le65_from_mx(mx_base),
    e65_clim = le65_from_mx(mx_clim),
    sd_base = sd65_from_mx(mx_base),
    sd_clim = sd65_from_mx(mx_clim))
}

stats_all <- rbindlist(stats_list)

cat("=== LE at 65 and LI (ages 65+) ===\n")
print(stats_all[year %in% decomp_years])
```

---

# Lloyd-Style Decomposition via `DemoDecomp::horiuchi()`

Following the Aburto et al. (2022) approach as applied by Lloyd et al. (2024), the Horiuchi method (continuous stepwise replacement, N=50) decomposes $\Delta = f(\text{pars}_\text{clim}) - f(\text{pars}_\text{base})$ into additive contributions from each element of the parameter vector. By passing a flattened vector of cause-specific mortality rates for ages 65+, the decomposition simultaneously attributes changes to age and cause.

```{r decompose-horiuchi}
n_causes <- length(temp_ranges) + 1
cause_names <- c(temp_labels, "rest")

decomp_list <- list()

for (yr in decomp_years) {
  yr_data <- analysis[year == yr & age >= cond_age]
  ages <- yr_data$age
  n_ages <- length(ages)

  mx_rest <- yr_data$deaths_baseline / yr_data$population
  mx_rest[is.nan(mx_rest) | is.infinite(mx_rest)] <- 1e-10

  pars1 <- rep(mx_rest, n_causes)
  names(pars1) <- paste(rep(ages, n_causes),
    rep(cause_names, each = n_ages), sep = "_")

  mx_cause_list <- lapply(temp_ranges, function(tr) {
    mx_c <- (yr_data$deaths_baseline + yr_data[[tr]]) / yr_data$population
    mx_c[is.nan(mx_c) | is.infinite(mx_c)] <- 1e-10
    mx_c
  })
  pars2 <- c(do.call(c, mx_cause_list), mx_rest)
  names(pars2) <- names(pars1)

  # LE decomposition
  cat(sprintf("  Decomposing e65 for %d...\n", yr))
  hor_le <- horiuchi(func = le65_cod_fun, pars1 = pars1, pars2 = pars2,
    ages = ages, n_causes = n_causes, N = 50)

  le_contrib <- matrix(hor_le, nrow = n_ages, ncol = n_causes)
  colnames(le_contrib) <- cause_names

  for (k in seq_len(n_causes)) {
    decomp_list[[length(decomp_list) + 1L]] <- data.table(
      year = yr, measure = "le", age = ages,
      temp_range = cause_names[k], contribution = le_contrib[, k])
  }

  # SD decomposition
  cat(sprintf("  Decomposing SD for %d...\n", yr))
  hor_sd <- horiuchi(func = sd65_cod_fun, pars1 = pars1, pars2 = pars2,
    ages = ages, n_causes = n_causes, N = 50)

  sd_contrib <- matrix(hor_sd, nrow = n_ages, ncol = n_causes)
  colnames(sd_contrib) <- cause_names

  for (k in seq_len(n_causes)) {
    decomp_list[[length(decomp_list) + 1L]] <- data.table(
      year = yr, measure = "sd", age = ages,
      temp_range = cause_names[k], contribution = sd_contrib[, k])
  }
}

decomp_all <- rbindlist(decomp_list)

check <- decomp_all[, .(sum_contrib = sum(contribution),
  total_diff = if (measure[1] == "le")
    stats_all[year == .BY$year, e65_clim - e65_base]
  else stats_all[year == .BY$year, sd_clim - sd_base]),
  by = .(year, measure)]
check[, diff := sum_contrib - total_diff]
cat("=== Additivity Check ===\n")
print(check)
```

---

# Results and Visualisation

## LE Decomposition by Age

```{r plot-le-decomp-age}
le_age <- decomp_all[measure == "le" & year %in% decomp_years,
  .(contribution = sum(contribution)), by = .(year, age)]

ggplot(le_age, aes(x = age, y = contribution, fill = factor(year))) +
  geom_col(position = "dodge", width = 1) +
  labs(title = "Age-Specific Contributions to Δe65 (Climate vs Baseline)",
    subtitle = sprintf("%s | %s", unique(analysis$city), unique(analysis$gcm)),
    x = "Age", y = "Contribution to Δe65 (years)", fill = "Year") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  theme_minimal(base_size = 12)
```

## LE Decomposition by Age and Temperature Range

```{r plot-le-decomp-temp}
ggplot(decomp_all[measure == "le" & year %in% decomp_years],
  aes(x = age, y = contribution, fill = temp_range)) +
  geom_col(width = 1) +
  facet_wrap(~ year, ncol = 1, scales = "free_y") +
  labs(title = "Δe65 by Age and Temperature Range (Horiuchi Decomposition)",
    subtitle = sprintf("%s | %s", unique(analysis$city), unique(analysis$gcm)),
    x = "Age", y = "Contribution to Δe65 (years)", fill = "Temperature range") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  theme_minimal(base_size = 12)
```

## Lifespan Inequality Decomposition

```{r plot-li-decomp}
ggplot(decomp_all[measure == "sd" & year %in% decomp_years],
  aes(x = age, y = contribution, fill = temp_range)) +
  geom_col(width = 1) +
  facet_wrap(~ year, ncol = 1, scales = "free_y") +
  labs(title = "ΔSD (65+) by Age and Temp Range (Horiuchi Decomposition)",
    subtitle = sprintf("%s | %s", unique(analysis$city), unique(analysis$gcm)),
    x = "Age", y = "Contribution to ΔSD (years)", fill = "Temperature range") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  theme_minimal(base_size = 12)
```

## Lifespan Inequality Time Series

```{r plot-li-timeseries}
li_long <- melt(stats_all[year %in% decomp_years],
  id.vars = "year", measure.vars = patterns("sd_"),
  variable.name = "scenario", value.name = "sd")
li_long[, scenario := fifelse(grepl("base", scenario), "Baseline", "Climate-adjusted")]

ggplot(li_long, aes(x = factor(year), y = sd, fill = scenario)) +
  geom_col(position = "dodge") +
  labs(title = "Lifespan Inequality (SD) Among People Aged 65+",
    subtitle = sprintf("%s | %s", unique(analysis$city), unique(analysis$ssp)),
    x = "Year", y = "SD of age at death (years)", fill = "Scenario") +
  scale_fill_manual(values = c("Baseline" = "steelblue",
    "Climate-adjusted" = "tomato")) +
  theme_minimal(base_size = 12)
```

## Summary by Age Group and Temperature Range

```{r cumulative-contributions}
age_groups_wide <- list(c(65, 74), c(75, 84), c(85, 100))
age_group_labels <- c("65-74", "75-84", "85+")

summary_list <- list()
for (i in seq_along(age_groups_wide)) {
  lo <- age_groups_wide[[i]][1]
  hi <- age_groups_wide[[i]][2]
  s <- decomp_all[age >= lo & age <= hi,
    .(contribution = sum(contribution)),
    by = .(year, measure, temp_range)]
  s[, age_group := age_group_labels[i]]
  summary_list[[length(summary_list) + 1L]] <- s
}

summary_decomp <- rbindlist(summary_list)
setcolorder(summary_decomp, c("year", "age_group", "measure", "temp_range", "contribution"))

cat("=== Summary: Δe65 by Age Group and Temp Range ===\n")
le_summary <- dcast(summary_decomp[measure == "le" & year %in% decomp_years],
  year + age_group ~ temp_range, value.var = "contribution", fill = 0)
print(le_summary)
```

---

# Export Results

```{r export}
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

fwrite(decomp_all[measure == "le"], output_le_decomp)
fwrite(decomp_all[measure == "sd"], output_li_decomp)
fwrite(stats_all, output_li_series)

cat("=== Files Written ===\n")
cat(sprintf("  LE decomposition (by age & temp): %s\n", output_le_decomp))
cat(sprintf("  LI decomposition (by age & temp): %s\n", output_li_decomp))
cat(sprintf("  Lifespan inequality time series: %s\n", output_li_series))
```

---

# Notebook Summary

This notebook:

1. computed period life tables for ages 65+ using the piecewise-exponential method (`px = exp(-mx * nx)`) — matching Lloyd/Aburto exactly,
2. computed lifespan inequality (SD of age at death) among people aged 65+,
3. applied `DemoDecomp::horiuchi()` (continuous stepwise replacement, N=50) to decompose Δe65 and ΔSD,
4. passed cause-specific mortality rates directly to horiuchi for simultaneous age × cause attribution,
5. produced summary tables showing how temperature-attributable mortality drives changes in LE and LI in the 65+ population.

---

# Session Information

```{r session-info}
sessionInfo()
```