---
title: "Temperature-attributable mortality, life expectancy, and lifespan inequality"
subtitle: "Projected impacts in Madrid under climate change (2020–2100)"
author: "Pipeline results"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(data.table); library(ggplot2); library(scales); library(ungroup)
source("../R/period_lifetable.R"); source("../R/utils.R"); source("../R/load_data.R")

OUT <- "../results/production"
city_code <- "ES001C"
```

# Attributable numbers

```{r an-load}
d <- fread(file.path(OUT, "ans_annual_all_cities.csv"),
  select = c("city_code", "gcm", "ssp", "age_group", "temp_range", "year", "an_est"))
gcm <- unique(d$gcm)[1]
mad <- d[city_code == city_code & gcm == gcm & ssp == "3"]
```

Total ANs 86M from 854 cities × 21 GCMs × 3 SSPs. For Madrid (`r gcm`, SSP3-7.0):

```{r an-totals}
tot <- mad[, .(AN = sum(an_est)), by = .(year, temp_range)]
tot_wide <- dcast(tot, year ~ temp_range, value.var = "AN")
tot_wide[, total := extreme_cold + moderate_cold + moderate_heat + extreme_heat]
```

```{r plot-ans, fig.height=4}
ggplot(tot, aes(x = year, y = AN, colour = temp_range)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Year", y = "Attributable deaths", colour = "Temperature range",
    title = "Annual attributable deaths — Madrid") +
  scale_y_continuous(labels = comma) +
  theme_minimal()
```

Annual ANs rise from `r round(tot_wide[year==2020, total], 0)` in 2020 to `r round(tot_wide[year==2099, total], 0)` in 2099. Heat-related deaths grow rapidly; cold-related deaths decline slightly as temperatures warm.

---

# Life expectancy at age 65

## Method

We reconstruct age-specific mortality rates for Madrid as:

$$ m(a,y) = m_\text{base}(a) \times \text{ratio}_\text{Eurostat}(a,y) $$

where $m_\text{base}(a)$ is Madrid's baseline mortality at age $a$ (from `city_results.csv`, PCLM-disaggregated to single years) and $\text{ratio}_\text{Eurostat}(a,y)$ is the national mortality improvement trend from Eurostat data.

Temperature-attributable ANs (by wide age group) are PCLM-disaggregated to single ages and converted to mortality rate increments:

$$ m_\text{clim}(a,y) = m(a,y) + \frac{\text{AN}_\text{total}(a,y)}{P(a)} $$

Life tables are constructed using the piecewise-constant hazard model (Aburto et al. 2022).

```{r le-compute}
# Madrid baseline mortality by single age (PCLM)
city_res <- fread("../data/city_results.csv")
mad_base <- city_res[URAU_CODE == city_code & agegroup %in% c("65-74","75-84","85+")]
age_starts <- c(65, 75, 85)
base_pop <- setNames(mad_base$agepop, mad_base$agegroup)
base_death <- setNames(mad_base$death, mad_base$agegroup)
pop_pclm <- fitted(pclm(x = age_starts, y = base_pop, nlast = 16, out.step = 1,
  control = list(lambda = NA, kr = 2, deg = 3)))
death_pclm <- fitted(pclm(x = age_starts, y = base_death, nlast = 16, out.step = 1,
  control = list(lambda = NA, kr = 2, deg = 3)))
mx_single <- as.numeric(death_pclm / pop_pclm)
mx_single[is.nan(mx_single)] <- 1e-10
ages <- 65:100
pop_single <- as.numeric(pop_pclm)

# Eurostat trend
es_mort <- load_eurostat_mortality("ES", sex = "T",
  age_min = 65, age_max = 100, year_min = 2020, year_max = 2099)

# PCLM disaggregation for ANs (wide groups -> single ages)
age_starts_wide <- c(20, 45, 65, 75, 85)
pclm_an <- function(v) {
  as.numeric(fitted(pclm(x = age_starts_wide, y = pmax(v, 0), nlast = 16,
    out.step = 1, control = list(lambda = NA, kr = 2, deg = 3))))
}

le_list <- list()
for (yr in 2020:2099) {
  an_yr <- mad[year == yr]
  an_mat <- dcast(an_yr, age_group ~ temp_range, value.var = "an_est", fill = 0)
  an_vec <- pclm_an(an_mat$extreme_cold + an_mat$moderate_cold +
    an_mat$moderate_heat + an_mat$extreme_heat)
  an_65 <- an_vec[46:81]  # ages 65-100

  mx_ratio <- es_mort[year == yr, mx_ratio]
  mx_proj <- mx_single * mx_ratio
  mx_proj[mx_proj <= 0] <- 1e-10

  # Baseline life table (no temperature)
  lt_base <- lifetable(mx = mx_proj, age = ages, ax = 0.5)

  # Climate-adjusted (add ANs as mortality increment)
  mx_an <- an_65 / pop_single
  lt_clim <- lifetable(mx = pmax(mx_proj + mx_an, 1e-10), age = ages, ax = 0.5)

  le_list[[length(le_list) + 1]] <- data.table(
    year = yr,
    e65_base = lt_base[age == 65, ex],
    e65_clim = lt_clim[age == 65, ex])
}
le_dt <- rbindlist(le_list)
```

```{r plot-le65, fig.height=4}
le_m <- melt(le_dt, id.vars = "year", variable.name = "scenario", value.name = "e65")
ggplot(le_m, aes(x = year, y = e65, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Year", y = "Remaining life expectancy at 65 (years)", colour = "Scenario",
    title = "Life expectancy at age 65 — Madrid") +
  scale_colour_manual(values = c("e65_base" = "steelblue", "e65_clim" = "tomato"),
    labels = c("Baseline (no temperature)", "Climate-adjusted")) +
  theme_minimal()
```

Under SSP3-7.0, $e_{65}$ shows a `r round(le_dt[year==2099, e65_clim - e65_base], 2)`-year difference between climate-adjusted and baseline by 2099, driven by rising heat-related mortality partially offset by declining cold. The baseline continues improving (falling mortality rates), but temperature-attributable deaths progressively erode those gains.

---

# Lifespan inequality (SD of age at death, ages 65+)

```{r sd-compute}
sd_list <- list()
for (yr in 2020:2099) {
  an_yr <- mad[year == yr]
  an_mat <- dcast(an_yr, age_group ~ temp_range, value.var = "an_est", fill = 0)
  an_vec <- pclm_an(an_mat$extreme_cold + an_mat$moderate_cold +
    an_mat$moderate_heat + an_mat$extreme_heat)
  an_65 <- an_vec[46:81]

  mx_ratio <- es_mort[year == yr, mx_ratio]
  mx_proj <- mx_single * mx_ratio
  mx_proj[mx_proj <= 0] <- 1e-10

  # Baseline LI
  nx <- c(rep(1, 35), Inf)
  lx_base <- cumprod(c(1, head(exp(-mx_proj * nx), -1)))
  dx_base <- c(-diff(lx_base), tail(lx_base, 1))
  e65_base <- le_dt[year == yr, e65_base]
  sd_base <- sqrt(sum(dx_base * (ages + 0.5 - e65_base)^2) / lx_base[1])

  # Climate-adjusted LI
  mx_c <- pmax(mx_proj + an_65 / pop_single, 1e-10)
  lx_c <- cumprod(c(1, head(exp(-mx_c * nx), -1)))
  dx_c <- c(-diff(lx_c), tail(lx_c, 1))
  e65_c <- le_dt[year == yr, e65_clim]
  sd_clim <- sqrt(sum(dx_c * (ages + 0.5 - e65_c)^2) / lx_c[1])

  sd_list[[length(sd_list) + 1]] <- data.table(year = yr, sd_base, sd_clim)
}
sd_dt <- rbindlist(sd_list)
```

```{r plot-sd65, fig.height=4}
sd_m <- melt(sd_dt, id.vars = "year", variable.name = "scenario", value.name = "sd")
ggplot(sd_m, aes(x = year, y = sd, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Year", y = "SD of age at death (years)", colour = "Scenario",
    title = "Lifespan inequality among people aged 65+ — Madrid") +
  scale_colour_manual(values = c("sd_base" = "steelblue", "sd_clim" = "tomato"),
    labels = c("Baseline (no temperature)", "Climate-adjusted")) +
  theme_minimal()
```

SD rises from `r round(sd_dt[year==2020, sd_clim], 2)` to `r round(sd_dt[year==2099, sd_clim], 2)` under the climate scenario (vs `r round(sd_dt[year==2099, sd_base], 2)` baseline). Temperature-related mortality increases lifespan inequality because it disproportionately affects older ages, widening the age-at-death distribution.

---

# Summary

```{r summary}
cat("Madrid (", gcm, ", SSP3-7.0): 2020 → 2099\n", sep = "")
cat(sprintf("  Annual attributable deaths: %s → %s (Δ = %+.0f)\n",
  format(round(tot_wide[year==2020, total]), big.mark = ","),
  format(round(tot_wide[year==2099, total]), big.mark = ","),
  tot_wide[year==2099, total] - tot_wide[year==2020, total]))
cat(sprintf("  Δ heat deaths (moderate + extreme): %+.0f\n",
  tot_wide[year==2099, moderate_heat + extreme_heat] -
  tot_wide[year==2020, moderate_heat + extreme_heat]))
cat(sprintf("  Δ cold deaths (moderate + extreme): %+.0f\n",
  tot_wide[year==2099, moderate_cold + extreme_cold] -
  tot_wide[year==2020, moderate_cold + extreme_cold]))
cat(sprintf("  e65 baseline: %.2f → %.2f (Δ = %+.2f y)\n",
  le_dt[year==2020, e65_base], le_dt[year==2099, e65_base],
  le_dt[year==2099, e65_base] - le_dt[year==2020, e65_base]))
cat(sprintf("  e65 climate:  %.2f → %.2f (Δ = %+.2f y)\n",
  le_dt[year==2020, e65_clim], le_dt[year==2099, e65_clim],
  le_dt[year==2099, e65_clim] - le_dt[year==2020, e65_clim]))
cat(sprintf("  Δe65 (climate − baseline): %.2f → %.2f\n",
  le_dt[year==2020, e65_clim - e65_base],
  le_dt[year==2099, e65_clim - e65_base]))
cat(sprintf("  SD baseline: %.2f → %.2f\n",
  sd_dt[year==2020, sd_base], sd_dt[year==2099, sd_base]))
cat(sprintf("  SD climate:  %.2f → %.2f\n",
  sd_dt[year==2020, sd_clim], sd_dt[year==2099, sd_clim]))
```

```{r session-info}
sessionInfo()
```