---
title: "Object 1: dataset for analysis (SSP3-7.0)"
subtitle: "Temperature-attributable deaths and population, 65+, 854 European cities"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include = FALSE}
# Knit from anywhere inside the repository: paths are relative to the repo root
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 5.5)
```

```{r packages}
library(data.table)
library(arrow)
library(ggplot2)

# Object 1 from pipeline/06_collect.R; set OBJ1 to use another file
# (e.g. the 10-capitals prototype VIG/data/260924_object1_10cities_ssp3_ensemble.parquet)
obj1_file <- Sys.getenv("OBJ1", "results/europe/collected/object1_dataset.parquet")
stopifnot(file.exists(obj1_file))

range_levels <- c("ExtrCold", "ModCold", "ModHeat", "ExtrHeat")
range_labels <- c(ExtrCold = "Extreme cold", ModCold = "Moderate cold", ModHeat = "Moderate heat", ExtrHeat = "Extreme heat")
range_colors <- c(ExtrCold = "#2166ac", ModCold = "#67a9cf", ModHeat = "#ef8a62", ExtrHeat = "#b2182b")
branch_labels <- c(with_cc = "With climate change", without_cc = "Without climate change")
branch_colors <- c(with_cc = "#b2182b", without_cc = "#2166ac")

o1 <- as.data.table(read_parquet(obj1_file))
# Country and region from city_results.csv if the file does not carry them
if (!all(c("country", "region") %in% names(o1))) {
  meta <- unique(fread("data/city_results.csv")[, .(city = URAU_CODE, city_name = LABEL, country = CNTR_CODE, region)])
  o1 <- merge(o1, meta, by = "city")
}
o1[, period := sprintf("%d-%d", year - (year - 2020) %% 5, year - (year - 2020) %% 5 + 4)]
```

File: ``r obj1_file``. Ensemble-mean ANs over the 19 GCMs, central ERF estimates. Deaths for the four temperature ranges are the attributable numbers (ANs); `rest` holds all other deaths.

# 1. What the table contains

One row per city × SSP × scenario × year × single age × cause (four temperature ranges and rest), with deaths and population. Country and region can be derived by summing cities; they are attached here for convenience (Masselot regions).

```{r structure}
o1[, .(rows = format(.N, big.mark = ","), cities = uniqueN(city), countries = uniqueN(country),
       regions = uniqueN(region), ssp = paste(unique(ssp), collapse = ","),
       years = paste(range(year), collapse = "-"), ages = paste(range(age), collapse = "-"))]
head(o1[order(city, scenario, year, age, cause)], 10)
```

## Coverage

```{r coverage}
all_cities <- unique(fread("data/city_results.csv")[, .(city = URAU_CODE, city_name = LABEL, country = CNTR_CODE, region)])
missing <- all_cities[!city %in% o1$city]
cov <- merge(all_cities[, .(total = .N), by = region], unique(o1[, .(city, region)])[, .(included = .N), by = region], by = "region")
cov
if (nrow(missing)) missing else "All cities included"
```

# 2. Basic checks

```{r checks}
pop_scen <- dcast(unique(o1[cause == "rest", .(city, scenario, year, age, pop)]), city + year + age ~ scenario, value.var = "pop")
rest_scen <- dcast(o1[cause == "rest", .(city, scenario, year, age, deaths)], city + year + age ~ scenario, value.var = "deaths")
data.table(
  check = c("No negative deaths", "No missing values",
            "Population identical with / without CC", "Rest deaths identical with / without CC (Simon 18 Sep rule)"),
  result = c(o1[, all(deaths >= 0)], !anyNA(o1[, .(deaths, pop)]),
             pop_scen[, max(abs(with_cc - without_cc))] < 1e-6,
             rest_scen[, max(abs(with_cc - without_cc))] < 1e-6)
)
```

# 3. Europe: attributable deaths by temperature range

Sum over all included cities, annual mean within each 5-year period.

```{r europe-an}
eu <- o1[cause %in% range_levels, .(AN = sum(deaths)), by = .(scenario, period, year, cause)][
  , .(AN = mean(AN)), by = .(scenario, period, cause)]
eu[, `:=`(mid = as.integer(substr(period, 1, 4)) + 2, cause = factor(cause, levels = range_levels))]
ggplot(eu, aes(mid, AN / 1000, colour = scenario)) +
  geom_line(linewidth = 0.8) + geom_point(size = 1) +
  facet_wrap(~cause, scales = "free_y", labeller = labeller(cause = range_labels)) +
  scale_colour_manual(values = branch_colors, labels = branch_labels, name = NULL) +
  labs(x = "5-year period (midpoint)", y = "Attributable deaths per year (thousands)",
       title = "Europe, 65+: temperature-attributable deaths, with vs without climate change") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")
```

```{r europe-an-table}
eu_tab <- dcast(eu[period %in% c("2020-2024", "2045-2049", "2070-2074", "2095-2099")],
                period + cause ~ scenario, value.var = "AN")
eu_tab[, `:=`(extra = with_cc - without_cc)]
eu_tab[, lapply(.SD, function(x) if (is.numeric(x)) round(x) else x)]
```

# 4. Extra deaths due to climate change, by region

With − without CC, annual mean over 2095–2099, per 100,000 people aged 65+.

```{r region}
end <- o1[period == "2095-2099"]
pop_reg <- unique(end[cause == "rest" & scenario == "with_cc", .(city, year, age, pop, region)])[
  , .(pop = sum(pop)), by = .(region, year)][, .(pop = mean(pop)), by = region]
reg <- dcast(end[cause %in% range_levels, .(AN = sum(deaths) / 5), by = .(region, cause, scenario)],
             region + cause ~ scenario, value.var = "AN")
reg <- merge(reg, pop_reg, by = "region")
reg[, `:=`(extra = with_cc - without_cc, extra_per_100k = 1e5 * (with_cc - without_cc) / pop,
           cause = factor(cause, levels = range_levels))]
ggplot(reg, aes(region, extra_per_100k, fill = cause)) +
  geom_col() + geom_hline(yintercept = 0, colour = "grey40") +
  scale_fill_manual(values = range_colors, labels = range_labels, name = NULL) +
  labs(x = NULL, y = "Extra deaths per 100,000 aged 65+ per year",
       title = "Change in temperature-attributable deaths due to climate change, 2095-2099",
       subtitle = "With - without climate change, SSP3-7.0") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")
dcast(reg, region ~ cause, value.var = "extra_per_100k")[, lapply(.SD, function(x) if (is.numeric(x)) round(x, 1) else x)]
```

# 5. Countries

Net extra deaths (all four ranges) per 100,000 aged 65+, 2095–2099.

```{r country, fig.height = 7}
pop_c <- unique(end[cause == "rest" & scenario == "with_cc", .(city, year, age, pop, country)])[
  , .(pop = sum(pop)), by = .(country, year)][, .(pop = mean(pop)), by = country]
ctry <- dcast(end[cause %in% range_levels, .(AN = sum(deaths) / 5), by = .(country, region, scenario)],
              country + region ~ scenario, value.var = "AN")
ctry <- merge(ctry, pop_c, by = "country")[, extra_per_100k := 1e5 * (with_cc - without_cc) / pop]
ctry[, n_cities := o1[, uniqueN(city), by = country][ctry, on = "country", V1]]
ggplot(ctry, aes(reorder(country, extra_per_100k), extra_per_100k, fill = region)) +
  geom_col() + coord_flip() + geom_hline(yintercept = 0, colour = "grey40") +
  labs(x = NULL, y = "Net extra deaths per 100,000 aged 65+ per year", fill = "Region",
       title = "Net change in temperature-attributable deaths due to climate change, 2095-2099") +
  theme_minimal(base_size = 10)
```

# 6. Age profile of the extra deaths (Europe)

Extra deaths (with − without CC) per 100,000 at each single age, around 2050 and 2090.

```{r age, fig.height = 5}
snap <- o1[period %in% c("2045-2049", "2050-2054", "2085-2089", "2090-2094") & cause %in% range_levels]
snap[, snapshot := ifelse(year < 2070, "2045-2054", "2085-2094")]
ag <- snap[, .(AN = sum(deaths), pop = sum(pop)), by = .(snapshot, scenario, age, cause)]
ag <- dcast(ag, snapshot + age + cause ~ scenario, value.var = c("AN", "pop"))
ag[, `:=`(rate = 1e5 * (AN_with_cc - AN_without_cc) / pop_with_cc, cause = factor(cause, levels = range_levels))]
ggplot(ag, aes(age, rate, colour = cause, linetype = snapshot)) +
  geom_hline(yintercept = 0, colour = "grey60") + geom_line(linewidth = 0.7) +
  scale_colour_manual(values = range_colors, labels = range_labels, name = NULL) +
  labs(x = "Age", y = "Extra deaths per 100,000 per year", linetype = NULL,
       title = "Europe: extra temperature-attributable deaths due to climate change, by age") +
  theme_minimal(base_size = 11)
```

# Notes

- These are attributable deaths only. LE65 / LI65+ for Europe, regions and countries need the pooled life tables and decompositions, which are the next step.
- Cities missing from the table (Section 1) are those whose run failed; they are excluded from every sum.
- 5-year periods start in 2020, as in the pipeline.
